---
title: "Rule of 3 Analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    df_print: kable
    keep_tex: false
fontsize: 12pt
geometry: margin=1in
always_allow_html: yes
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{threeparttable}
params:
  tabset: 'Sample Size'
  incidence_rate: 100
  sample_size: 230
  power: 0.8
  discon: 0.10
  adj_n: 250
  progress: NA
---

```{r, include=FALSE}
options(tinytex.verbose = F)
```

```{r setup, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pwr)
library(binom)
library(kableExtra)
```

## About this analysis
This analysis performs power and sample size calculations for single proportion hypothesis testing using Cohen's arcsine transformation method. This statistical approach tests whether an observed event rate differs significantly from a null hypothesis (typically zero events or a specified reference rate).

The method uses the arcsine transformation of proportions, which provides accurate power calculations across a wide range of event rates. This is the standard approach recommended in Cohen (1988) "Statistical Power Analysis for the Behavioral Sciences" and is implemented using the pwr.p.test function in R.

This approach is particularly useful for studies examining rare event data in post-marketing surveillance (PMS) studies, safety monitoring in clinical trials, and quality control applications where detecting deviations from expected event rates is critical.

Reference: Cohen J. Statistical Power Analysis for the Behavioral Sciences. 2nd ed. Routledge; 1988.

## Results of this analysis
Based on the Binominal distribution and a true event incidence rate of 1 in `r format(params$incidence_rate, digits=2, nsmall=2)` (or `r format(1/params$incidence_rate, digits=2, nsmall=2)`%), `r format(ceiling(params$sample_size), digits=1, nsmall=0)` participants would be needed to observe at least one event with `r format(params$power*100, digits=2, nsmall=0)`% probability. Accounting for a possible withdrawal or discontinuation rate of `r format(params$discon*100, digits=1)`%, the target number of participants is set as `r format(ceiling((params$sample_size * (1+params$discon))), digits=1)`.

\newpage
### Figure 1. Estimated power for the given conditions at different sample sizes.
```{r figure1, echo=FALSE}
# Increment the progress bar, and update the detail text.
params$progress$inc(1/3, detail = "Making Figure 2")

p.out <- if(params$tabset == 'Power'){
    pwr.p.test(sig.level=0.05, power=NULL, h = ES.h(1/params$incidence_rate, 0), alt="greater", n = params$sample_size)
  } else {
    pwr.p.test(sig.level=0.05, power=params$power, h = ES.h(1/params$incidence_rate, 0), alt="greater", n = NULL)
  }
      
plot(p.out)
```

\newpage
### Table 1. Event rate accuracy with 95% confidence interval limits
If `r format(ceiling(params$sample_size), digits=1, nsmall=0)` participants are included, the event rate would be estimated to an accuracy shown in the table below:


```{r table1, echo=FALSE, results='asis'}
# Increment the progress bar, and update the detail text.
params$progress$inc(1/4, detail = "Creating Table 1")

sequence <- unique(c(seq(0, 5), seq(10, 25, by=5), seq(50, min(round(params$sample_size, 0), 1000), by=50), seq(min(round(params$sample_size, 0), 1000), min(round(params$sample_size, 0), 10000), by=1000)))
bb <- lapply(sequence, function(n) {
  binom.confint(n, params$sample_size, conf.level = 0.95, methods = "exact")
})

table <- do.call(rbind, bb)
table$length <- table$upper - table$lower
var <- c("mean", "lower", "upper", "length")
for (i in var) {
  table[, i] <- round(table[, i]*100, 1)
}

table <- table[, c(2,4:7)]
kable(table, align='lrrrr', 
      col.names = c('Number of Events Observed', "Event Rate$^1$", "Lower Limit$^2$", "Upper Limit$^2$",'Length'), 
      format = "latex", booktabs=T, escape=F) %>%
  kable_styling(full_width = FALSE, position="center", fixed_thead = T) %>%
  add_header_above(c(" ", " ", "95% Confidence Interval" = 2, " ")) %>%
  footnote(number = c("Event rate (%) is estimated as a crude rate, defined as the number of participants exposed and experiencing the event of interest divided by the total number of participants. ",
                      "Confidence Interval (%) based on exact Clopper-Pearson exact method for one proportion.")
           , threeparttable = T)
params$progress$inc(1/5, detail = "Writing PDF File (May take a few minutes)")
```