---
title: "Design Effects in Clustered Studies: ICC Selection and Sample Size Inflation"
author: "Power Analysis Tool Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design Effects in Clustered Studies: ICC Selection and Sample Size Inflation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Many real-world studies involve **clustered** or **hierarchical** data where observations are not independent:

- **Patients within hospitals** or clinics
- **Students within schools** or classrooms
- **Individuals within households** or geographic regions
- **Repeated measures within participants**

Clustering reduces effective sample size because observations within the same cluster tend to be more similar than observations from different clusters. This similarity is quantified by the **Intraclass Correlation Coefficient (ICC)**.

This vignette explains:

1. **What is the ICC** and how it measures clustering
2. **Design effect (DE)** and its relationship to ICC and cluster size
3. **How to select appropriate ICC values** from literature or pilot data
4. **Sample size inflation** calculations for clustered designs
5. **Practical examples** across different study types

## Understanding the ICC

### Definition

The **Intraclass Correlation Coefficient (ICC)** measures the proportion of total variance in an outcome that is attributable to **between-cluster** variation:

$$
ICC = \frac{\sigma^2_{between}}{\sigma^2_{between} + \sigma^2_{within}}
$$

where:
- $\sigma^2_{between}$ = variance between clusters
- $\sigma^2_{within}$ = variance within clusters

### Interpretation

**ICC = 0**: No clustering effect
- Observations within a cluster are no more similar than observations from different clusters
- Cluster membership is irrelevant
- Standard sample size formulas apply (no inflation needed)

**ICC = 0.05**: Moderate clustering
- 5% of variance is due to cluster-level differences
- 95% of variance is within-cluster (individual-level)
- **Most common in health research**

**ICC = 0.30**: Strong clustering
- 30% of variance is at cluster level
- Observations within a cluster are highly similar
- Large sample size inflation required

**ICC = 1.0**: Perfect clustering
- All variance is between clusters
- Observations within a cluster are identical
- Cluster is the unit of analysis (not individuals)

### Analogy

Imagine measuring blood pressure in hospitals:

- **ICC = 0**: Hospital doesn't matter; patients' BPs vary regardless of hospital
- **ICC = 0.05**: Hospitals have slightly different patient populations or protocols; 5% of BP variation is due to which hospital you're at
- **ICC = 0.30**: Hospitals serve very different populations (e.g., cardiac specialty vs general); hospital explains 30% of BP variation
- **ICC = 1.0**: All patients in a hospital have identical BP (impossible in practice)

## Design Effect

### Definition

The **Design Effect (DE)** quantifies the **loss of statistical efficiency** due to clustering. It is the factor by which sample size must be inflated to maintain the same power as an independent-sample design.

$$
DE = 1 + (m - 1) \times ICC
$$

where:
- **m** = average cluster size (number of participants per cluster)
- **ICC** = intraclass correlation coefficient

### Properties

**When ICC = 0**: DE = 1 (no inflation needed)

**When ICC > 0 and m > 1**: DE > 1 (sample size inflation required)

**Larger clusters amplify the design effect**: Doubling cluster size doubles the design effect when ICC is substantial

**Larger ICC amplifies the design effect**: Doubling ICC doubles the contribution to design effect

### Examples

```{r de-examples, echo=FALSE}
de_examples <- data.frame(
  Cluster_Size = c(10, 25, 50, 100, 10, 25, 50, 100, 10, 25, 50, 100),
  ICC = c(rep(0.01, 4), rep(0.05, 4), rep(0.10, 4)),
  Design_Effect = c(
    1 + (10-1)*0.01, 1 + (25-1)*0.01, 1 + (50-1)*0.01, 1 + (100-1)*0.01,
    1 + (10-1)*0.05, 1 + (25-1)*0.05, 1 + (50-1)*0.05, 1 + (100-1)*0.05,
    1 + (10-1)*0.10, 1 + (25-1)*0.10, 1 + (50-1)*0.10, 1 + (100-1)*0.10
  ),
  stringsAsFactors = FALSE
)

library(knitr)
kable(de_examples, caption = "Design Effect by Cluster Size and ICC", digits = 2)
```

**Observations**:

1. **Small ICC (0.01)**: Even with 100 per cluster, DE = 2.0 (modest inflation)
2. **Moderate ICC (0.05)**: With 50 per cluster, DE = 3.45 (substantial inflation)
3. **Large ICC (0.10)**: With 100 per cluster, DE = 10.9 (severe inflation)

**Key insight**: Large clusters with moderate-to-high ICC can dramatically inflate sample size requirements.

## Sample Size Inflation for Clustered Designs

### Basic Formula

If an unclustered study requires N participants for adequate power:

$$
N_{clustered} = N_{unclustered} \times DE
$$

### Effective Sample Size

Conversely, if you have N participants in a clustered design, the effective sample size (equivalent independent observations) is:

$$
N_{effective} = \frac{N_{total}}{DE}
$$

### Number of Clusters

Given total sample size and average cluster size:

$$
K = \frac{N_{total}}{m}
$$

where K is the number of clusters needed.

### Practical Example

**Scenario**: Cluster-randomized trial of a quality improvement intervention across hospitals

- Outcome: 30-day readmission rate
- Unclustered sample size needed: **N = 400** (from standard power calculation)
- Average hospital size: **m = 50 patients** per hospital
- ICC for readmission: **0.05** (typical for clinical outcomes)

**Calculations**:

```{r example-calc, eval=FALSE}
library(PowerAnalysisTool)

# Step 1: Calculate design effect
de <- calc_design_effect(cluster_size = 50, icc = 0.05)
# Result: DE = 1 + (50 - 1) × 0.05 = 3.45

# Step 2: Inflate sample size
n_clustered <- calc_clustered_n(n_unclustered = 400, design_effect = de)
# Result: N = 400 × 3.45 = 1,380 participants

# Step 3: Calculate number of clusters needed
k <- calc_n_clusters(n_total = 1380, cluster_size = 50)
# Result: K = 1,380 / 50 = 28 hospitals

# Interpretation
interp <- format_design_effect_interpretation(
  design_effect = de,
  n_unclustered = 400,
  cluster_size = 50,
  icc = 0.05
)
print(interp)
```

**Output**:
```
The design effect of 3.45 indicates a strong clustering impact. Clustering
inflates the required sample size by 245% (from 400 to 1,380 participants,
an increase of 980). With an average cluster size of 50 and ICC of 0.050,
you will need approximately 28 clusters.
```

**Result**: Instead of 400 independent patients, need **1,380 patients across 28 hospitals** to maintain 80% power.

## Selecting Appropriate ICC Values

Choosing the right ICC is critical for accurate sample size estimation. Three approaches:

### 1. Pilot Study or Preliminary Data

**Best approach** if you have access to similar data.

**Method**: Fit a mixed-effects model (random intercept for cluster) and extract the ICC:

```{r icc-pilot, eval=FALSE}
# Example in R using lme4
library(lme4)

# Fit random intercept model: outcome ~ fixed effects + (1 | cluster)
model <- lmer(blood_pressure ~ age + sex + treatment + (1 | hospital_id), data = pilot_data)

# Extract variance components
vc <- as.data.frame(VarCorr(model))
var_between <- vc$vcov[vc$grp == "hospital_id"]
var_within <- vc$vcov[vc$grp == "Residual"]

# Calculate ICC
icc_estimate <- var_between / (var_between + var_within)
cat("Estimated ICC:", round(icc_estimate, 3))
```

### 2. Literature-Based Typical Values

When pilot data are unavailable, use published ICC estimates from meta-analyses.

**Typical ICC ranges by domain** (Donner & Klar 2000; Campbell et al. 2005):

```{r typical-icc, echo=FALSE}
icc_table <- data.frame(
  Domain = c("Behavioral outcomes",
             "Clinical/physiological outcomes",
             "Process measures",
             "General practice level",
             "Mortality outcomes",
             "Hospital characteristics",
             "Classroom-level education"),
  Typical_ICC = c("0.01 - 0.05",
                  "0.01 - 0.10",
                  "0.10 - 0.30",
                  "0.017 (median)",
                  "0.005 - 0.015",
                  "0.05 - 0.20",
                  "0.10 - 0.25"),
  Recommended = c("0.025",
                  "0.05",
                  "0.20",
                  "0.017",
                  "0.01",
                  "0.10",
                  "0.15"),
  Example = c("Smoking cessation, exercise adherence",
              "Blood pressure, HbA1c, weight",
              "Guideline adherence, prescribing patterns",
              "UK GP practices (various outcomes)",
              "30-day mortality, in-hospital death",
              "Length of stay, complications",
              "Test scores, attendance"),
  stringsAsFactors = FALSE
)

kable(icc_table, caption = "Typical ICC Values by Domain and Outcome Type")
```

**Using the tool's built-in values**:

```{r builtin-icc, eval=FALSE}
# Get typical ICC for clinical outcomes
icc_clinical <- get_typical_icc("clinical")
# Returns: 0.05

# Get typical ICC for behavioral outcomes
icc_behavioral <- get_typical_icc("behavioral")
# Returns: 0.025

# Get typical ICC for process measures
icc_process <- get_typical_icc("process")
# Returns: 0.20

# Get typical ICC for GP practices
icc_gp <- get_typical_icc("gp")
# Returns: 0.017
```

### 3. Sensitivity Analysis

**Always conduct sensitivity analysis** to assess robustness to ICC assumptions.

Test best-case, expected, and worst-case scenarios:

```{r sensitivity-icc, eval=FALSE}
# Expected ICC: 0.05
de_expected <- calc_design_effect(cluster_size = 40, icc = 0.05)
n_expected <- calc_clustered_n(n_unclustered = 300, design_effect = de_expected)

# Best case (lower ICC): 0.02
de_best <- calc_design_effect(cluster_size = 40, icc = 0.02)
n_best <- calc_clustered_n(n_unclustered = 300, design_effect = de_best)

# Worst case (higher ICC): 0.10
de_worst <- calc_design_effect(cluster_size = 40, icc = 0.10)
n_worst <- calc_clustered_n(n_unclustered = 300, design_effect = de_worst)

cat("Best case (ICC=0.02):", n_best, "participants\n")
cat("Expected (ICC=0.05):", n_expected, "participants\n")
cat("Worst case (ICC=0.10):", n_worst, "participants\n")
```

**Output**:
```
Best case (ICC=0.02): 624 participants
Expected (ICC=0.05): 888 participants
Worst case (ICC=0.10): 1,488 participants
```

**Decision**: Plan for worst-case (1,488) or expected (888) based on risk tolerance and feasibility.

## Study Design Considerations

### Trade-off: Cluster Size vs Number of Clusters

For a fixed ICC and total sample size, you can trade off:

- **Many small clusters** (lower DE, better efficiency)
- **Fewer large clusters** (higher DE, worse efficiency)

**Example**: N = 1,000 participants, ICC = 0.05

```{r tradeoff, echo=FALSE}
tradeoff_data <- data.frame(
  Scenario = c("A: Many small clusters", "B: Moderate", "C: Moderate", "D: Few large clusters"),
  N_Clusters = c(100, 50, 25, 10),
  Cluster_Size = c(10, 20, 40, 100),
  Design_Effect = c(
    1 + (10-1)*0.05,
    1 + (20-1)*0.05,
    1 + (40-1)*0.05,
    1 + (100-1)*0.05
  ),
  Effective_N = c(
    1000 / (1 + (10-1)*0.05),
    1000 / (1 + (20-1)*0.05),
    1000 / (1 + (40-1)*0.05),
    1000 / (1 + (100-1)*0.05)
  ),
  stringsAsFactors = FALSE
)

kable(tradeoff_data, caption = "Trade-off Between Cluster Size and Number of Clusters", digits = 1)
```

**Observations**:

- **Scenario A** (100 clusters of 10): Effective N = 690 (best efficiency)
- **Scenario D** (10 clusters of 100): Effective N = 172 (worst efficiency)

**Recommendation**: **Recruit more clusters with fewer participants per cluster** when feasible. This is more efficient than fewer large clusters.

**Practical constraint**: Recruiting clusters (e.g., hospitals) is often difficult and expensive, so there's a practical limit.

### Minimum Number of Clusters

**Rule of thumb**: **At least 10-15 clusters per treatment arm** (20-30 total for two-arm trial)

**Why?**
- Variance estimation for cluster-level effects requires adequate degrees of freedom
- Fewer clusters → unreliable variance estimates → inflated Type I error
- Random effects models struggle with < 10 clusters

```{r validation-example, eval=FALSE}
# Validate clustering parameters
validation <- validate_clustering_params(
  n_clusters = 8,
  cluster_size = 50,
  icc = 0.05
)

if (!validation$valid) {
  cat("ERRORS detected:\n")
  print(validation$messages[grepl("ERROR", validation$messages)])
}

if (any(grepl("WARNING", validation$messages))) {
  cat("\nWARNINGS:\n")
  print(validation$messages[grepl("WARNING", validation$messages)])
}
```

**Output**:
```
WARNINGS:
[1] "WARNING: Only 8 clusters may provide unreliable variance estimates.
     Consider 10-15+ clusters."
```

**Interpretation**: 8 clusters is technically feasible but statistically risky. Aim for 12-15+ clusters if possible.

## Case Study 1: Cluster-Randomized Trial in Primary Care

### Scenario

A primary care practice network wants to test a diabetes management intervention. Practices will be randomized to receive the intervention or usual care.

**Design parameters**:
- **Outcome**: HbA1c reduction (continuous)
- **Unclustered N**: 220 (from independent-sample t-test for 80% power, d = 0.40)
- **Cluster type**: GP practices
- **Average practice size**: 30 diabetes patients per practice
- **ICC**: 0.017 (median ICC for GP practice-level outcomes)

### Calculations

```{r case1-calc, eval=FALSE}
# Step 1: Design effect
de <- calc_design_effect(cluster_size = 30, icc = 0.017)
# Result: DE = 1 + (30 - 1) × 0.017 = 1.493

# Step 2: Inflated sample size
n_clustered <- calc_clustered_n(n_unclustered = 220, design_effect = de)
# Result: N = 220 × 1.493 = 329 patients

# Step 3: Number of practices
k <- calc_n_clusters(n_total = 329, cluster_size = 30)
# Result: K = 329 / 30 = 11 practices

# Interpretation
cat("Need:", n_clustered, "patients across", k, "practices\n")
cat("Per arm:", ceiling(k / 2), "practices (assuming equal allocation)\n")
```

**Output**:
```
Need: 329 patients across 11 practices
Per arm: 6 practices (assuming equal allocation)
```

### Sensitivity Analysis

```{r case1-sens, eval=FALSE}
# Worst case: ICC = 0.05 (clinical outcome, upper range)
de_worst <- calc_design_effect(cluster_size = 30, icc = 0.05)
n_worst <- calc_clustered_n(n_unclustered = 220, design_effect = de_worst)
k_worst <- calc_n_clusters(n_total = n_worst, cluster_size = 30)

cat("Worst case (ICC=0.05):", n_worst, "patients across", k_worst, "practices\n")
```

**Output**:
```
Worst case (ICC=0.05): 550 patients across 19 practices
```

### Decision

- **Expected scenario (ICC = 0.017)**: 11 practices (6 per arm)
- **Worst case (ICC = 0.05)**: 19 practices (10 per arm)

**Recommendation**: Recruit **20 practices** (10 per arm) to be conservative and ensure adequate cluster count even under higher ICC.

## Case Study 2: Hospital-Level Quality Improvement

### Scenario

A hospital network wants to reduce 30-day readmission rates through a quality improvement intervention. Hospitals will be randomized.

**Design parameters**:
- **Outcome**: 30-day readmission (binary, proportion)
- **Unclustered N**: 500 (from two-proportion test for 80% power, p1 = 0.20, p2 = 0.15)
- **Cluster type**: Hospitals
- **Average hospital size**: 100 patients per hospital
- **ICC**: 0.05 (clinical outcome, typical value)

### Problem: Very Large Design Effect

```{r case2-problem, eval=FALSE}
de <- calc_design_effect(cluster_size = 100, icc = 0.05)
# Result: DE = 1 + (100 - 1) × 0.05 = 5.95

n_clustered <- calc_clustered_n(n_unclustered = 500, design_effect = de)
# Result: N = 500 × 5.95 = 2,975 patients

k <- calc_n_clusters(n_total = 2975, cluster_size = 100)
# Result: K = 2,975 / 100 = 30 hospitals
```

**Issue**: Need 30 hospitals with 100 patients each (2,975 total) — likely infeasible.

### Solution: Reduce Cluster Size

Instead of 100 patients per hospital, sample **50 patients per hospital**:

```{r case2-solution, eval=FALSE}
de_reduced <- calc_design_effect(cluster_size = 50, icc = 0.05)
# Result: DE = 1 + (50 - 1) × 0.05 = 3.45

n_clustered_reduced <- calc_clustered_n(n_unclustered = 500, design_effect = de_reduced)
# Result: N = 500 × 3.45 = 1,725 patients

k_reduced <- calc_n_clusters(n_total = 1725, cluster_size = 50)
# Result: K = 1,725 / 50 = 35 hospitals
```

**Trade-off**:
- **Smaller cluster size (50)**: Need 35 hospitals but only 1,725 patients
- **Larger cluster size (100)**: Need 30 hospitals but 2,975 patients

**Recommendation**: Use **50 patients per hospital** to reduce total sample size and make recruitment more feasible. The additional 5 hospitals (35 vs 30) may be easier to recruit than an extra 1,250 patients.

### Alternative: Recruit More Hospitals, Smaller Clusters

```{r case2-alternative, eval=FALSE}
# Even smaller: 25 patients per hospital
de_small <- calc_design_effect(cluster_size = 25, icc = 0.05)
n_small <- calc_clustered_n(n_unclustered = 500, design_effect = de_small)
k_small <- calc_n_clusters(n_total = n_small, cluster_size = 25)

cat("Option C (25 per hospital):", n_small, "patients across", k_small, "hospitals\n")
```

**Output**:
```
Option C (25 per hospital): 1,100 patients across 44 hospitals
```

**Final comparison**:

| Option | Cluster Size | Design Effect | Total N | Hospitals | Feasibility |
|--------|--------------|---------------|---------|-----------|-------------|
| A | 100 | 5.95 | 2,975 | 30 | Low (too many patients) |
| B | 50 | 3.45 | 1,725 | 35 | Moderate |
| C | 25 | 2.20 | 1,100 | 44 | High (most efficient) |

**Decision**: If 44 hospitals are recruitable, **Option C is optimal** (smallest total sample size and highest efficiency).

## Advanced Topics

### Unequal Cluster Sizes

When cluster sizes vary, use the **average cluster size** (m_bar) or, more conservatively, the **maximum anticipated cluster size** for design effect calculations.

**Example**:
- Hospital A: 30 patients
- Hospital B: 50 patients
- Hospital C: 70 patients

Average: m_bar = (30 + 50 + 70) / 3 = 50

```{r unequal-clusters, eval=FALSE}
# Use average cluster size
de_avg <- calc_design_effect(cluster_size = 50, icc = 0.05)

# Or use maximum for conservative estimate
de_max <- calc_design_effect(cluster_size = 70, icc = 0.05)
```

### Varying ICC Across Outcomes

Different outcomes within the same study may have different ICCs:

- **Primary outcome**: Use published ICC or pilot estimate
- **Secondary outcomes**: Consider sensitivity analysis with different ICCs

### Cluster-Randomized vs Individually-Randomized Designs

**When to use cluster randomization:**
- Intervention is delivered at cluster level (e.g., training all staff in a hospital)
- Risk of contamination if individuals within a cluster receive different treatments
- Logistical reasons (easier to randomize clusters than individuals)

**Cost**:
- Cluster randomization requires larger sample sizes due to design effect
- Only beneficial when ICC is unavoidable or intervention logistics demand it

**Alternative**: If individual randomization is feasible, it's **statistically more efficient** (DE = 1, no inflation).

### Adjusting for Clustering in Analysis

**Important**: Planning for clustered data requires:

1. **Design stage**: Inflate sample size using design effect
2. **Analysis stage**: Use appropriate statistical methods:
   - **Mixed-effects models** (random intercept for cluster)
   - **Generalized Estimating Equations (GEE)**
   - **Cluster-robust standard errors**

Failing to account for clustering in analysis invalidates p-values and confidence intervals (typically anti-conservative, inflated Type I error).

## Practical Workflow Summary

### Step 1: Calculate Unclustered Sample Size

Use standard sample size formulas (t-test, two-proportion test, etc.) assuming independent observations.

```{r workflow-step1, eval=FALSE}
# Example: Two-sample t-test
library(pwr)
result <- pwr.t.test(d = 0.40, sig.level = 0.05, power = 0.80, type = "two.sample")
n_unclustered <- ceiling(result$n * 2)  # Total for both groups
```

### Step 2: Determine ICC

From pilot data, literature, or tool's typical values:

```{r workflow-step2, eval=FALSE}
# Option A: Pilot data (fit mixed model, extract ICC)
# Option B: Literature values
icc <- get_typical_icc("clinical")  # Returns 0.05
# Option C: Conservative assumption (e.g., upper bound from lit)
```

### Step 3: Estimate Average Cluster Size

Based on feasibility and cluster characteristics:

```{r workflow-step3, eval=FALSE}
cluster_size <- 40  # Example: 40 patients per practice
```

### Step 4: Calculate Design Effect

```{r workflow-step4, eval=FALSE}
de <- calc_design_effect(cluster_size = cluster_size, icc = icc)
```

### Step 5: Inflate Sample Size

```{r workflow-step5, eval=FALSE}
n_clustered <- calc_clustered_n(n_unclustered = n_unclustered, design_effect = de)
```

### Step 6: Determine Number of Clusters

```{r workflow-step6, eval=FALSE}
k <- calc_n_clusters(n_total = n_clustered, cluster_size = cluster_size)

# Check if adequate (>=10 per arm for 2-arm trial)
k_per_arm <- ceiling(k / 2)
if (k_per_arm < 10) {
  cat("WARNING: Only", k_per_arm, "clusters per arm. Aim for 10-15+ per arm.\n")
}
```

### Step 7: Validate Parameters

```{r workflow-step7, eval=FALSE}
validation <- validate_clustering_params(
  n_clusters = k,
  cluster_size = cluster_size,
  icc = icc
)

if (!validation$valid) {
  stop("Clustering parameters are invalid. Review errors.")
}

if (length(validation$messages) > 0) {
  cat("Validation messages:\n")
  print(validation$messages)
}
```

### Step 8: Sensitivity Analysis

```{r workflow-step8, eval=FALSE}
# Test worst-case scenario
icc_worst <- icc * 2  # Double the ICC
de_worst <- calc_design_effect(cluster_size = cluster_size, icc = icc_worst)
n_worst <- calc_clustered_n(n_unclustered = n_unclustered, design_effect = de_worst)
k_worst <- calc_n_clusters(n_total = n_worst, cluster_size = cluster_size)

cat("Expected scenario:", k, "clusters,", n_clustered, "participants\n")
cat("Worst case:", k_worst, "clusters,", n_worst, "participants\n")
```

### Step 9: Make Final Decision

Balance statistical requirements with feasibility:

- Use expected scenario if confident in ICC estimate
- Use worst-case if ICC is uncertain or recruitment is feasible
- Consider reducing cluster size if design effect is prohibitively large

## Summary

**Key takeaways**:

1. **Clustering reduces effective sample size** via the design effect: DE = 1 + (m - 1) × ICC
2. **Inflate sample size by the design effect** to maintain power: N_clustered = N_unclustered × DE
3. **Use literature-based typical ICC values** when pilot data are unavailable:
   - Behavioral outcomes: 0.025
   - Clinical outcomes: 0.05
   - Process measures: 0.20
   - GP practice level: 0.017
4. **Many small clusters are more efficient** than few large clusters
5. **Aim for 10-15+ clusters per arm** for reliable inference
6. **Conduct sensitivity analysis** to assess robustness to ICC assumptions
7. **Account for clustering in analysis** using mixed models or GEE

**Recommended resources**:

- Donner A, Klar N (2000). *Design and Analysis of Cluster Randomization Trials in Health Research*. Arnold Publishers.
- Campbell MK, Thomson S, Ramsay CR, et al. (2004). Sample size calculator for cluster randomized trials. *Computers in Biology and Medicine* 34(2):113-125.
- Eldridge SM, Ashby D, Kerry S (2006). Sample size for cluster randomized trials: effect of coefficient of variation of cluster size and analysis method. *International Journal of Epidemiology* 35(5):1292-1300.

## Further Reading

- See `vignette("propensity-score-calculations")` for handling clustering in observational studies
- See `vignette("missing-data-adjustments")` for combining clustering and missing data adjustments
- See the Clustering module in all tabs of the Power Analysis Tool for interactive calculations
