---
title: "Interactive Power Curves: Visualizing Trade-offs in Study Design"
author: "Power Analysis Tool Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive Power Curves: Visualizing Trade-offs in Study Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Traditional power calculations produce a **single number** (required sample size for a specific effect size). However, study planning involves **trade-offs**:

- What if the true effect is larger or smaller than assumed?
- How much does power increase if we recruit 10% more participants?
- What effect sizes are we well-powered vs underpowered to detect?

**Interactive power curves** visualize these trade-offs by showing **power as a function of effect size** (for fixed sample size) or **power as a function of sample size** (for fixed effect size).

This vignette explains:

1. **What are power curves** and why they're more informative than single-point calculations
2. **How to interpret power curves** for different study designs
3. **Common patterns** and what they reveal about your study
4. **How to use power curves** for sensitivity analysis and decision-making
5. **Practical examples** with real study scenarios

## What Are Power Curves?

### Definition

A **power curve** is a graphical representation of statistical power across a **range of values**:

- **Type 1: Power vs Effect Size** (fixed N)
  - X-axis: Effect size (e.g., Cohen's d, relative risk, hazard ratio)
  - Y-axis: Statistical power (0 to 1)
  - Shows: What effects can we detect with our sample size?

- **Type 2: Power vs Sample Size** (fixed effect)
  - X-axis: Sample size (N)
  - Y-axis: Statistical power (0 to 1)
  - Shows: How does power increase as we add participants?

### Why Power Curves Matter

**Problem with single-point calculations**:

Traditional power calculation:
> "To detect an effect of d = 0.50 with 80% power, you need N = 128."

**But**:
- What if the true effect is d = 0.40? (Underpowered)
- What if the true effect is d = 0.60? (Overpowered; could use smaller N)
- What if we can only recruit N = 100? (What power do we have?)

**Solution: Power curves answer all of these questions at once.**

### Anatomy of a Power Curve

```{r power-curve-anatomy, echo=FALSE, eval=FALSE}
# Example power curve (conceptual)
library(ggplot2)
library(pwr)

# Generate power curve for two-sample t-test
n <- 64  # per group
effect_sizes <- seq(0, 1, by = 0.01)
power_values <- sapply(effect_sizes, function(d) {
  if (d == 0) return(0.05)  # Type I error rate at null
  result <- pwr.t.test(n = n, d = d, sig.level = 0.05, type = "two.sample", alternative = "two.sided")
  return(result$power)
})

df <- data.frame(Effect_Size = effect_sizes, Power = power_values)

ggplot(df, aes(x = Effect_Size, y = Power)) +
  geom_line(linewidth = 1.2, color = "#1f77b4") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0.50, linetype = "dashed", color = "darkgreen") +
  annotate("text", x = 0.55, y = 0.82, label = "80% Power Threshold", color = "red", hjust = 0) +
  annotate("text", x = 0.52, y = 0.15, label = "Target Effect\n(d = 0.50)", color = "darkgreen", hjust = 0) +
  annotate("point", x = 0.50, y = 0.80, color = "purple", size = 3) +
  labs(
    title = "Power Curve: Two-Sample t-test (N = 64 per group)",
    x = "Effect Size (Cohen's d)",
    y = "Statistical Power",
    caption = "Dashed red line = 80% power threshold; Dashed green line = Target effect size"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

**Key features**:

1. **Sigmoid shape**: Power increases slowly for small effects, rapidly in middle range, then plateaus near 1.0 for large effects

2. **Alpha level (Type I error)**: At effect size = 0 (null hypothesis), power equals α (typically 0.05)

3. **Target point**: Intersection of target effect size and desired power (e.g., d = 0.50, power = 0.80)

4. **Minimal Detectable Effect**: Effect size where power curve crosses 0.80 threshold

5. **Underpowered region**: Left of MDE (power < 0.80)

6. **Well-powered region**: Right of MDE (power ≥ 0.80)

## Interpreting Power Curves for Different Designs

### Two-Sample Comparison (Continuous Outcome)

**Scenario**: Comparing two groups on a continuous outcome (e.g., blood pressure).

**Power curve interpretation**:

```{r two-sample-interpretation, echo=FALSE}
interpretation <- data.frame(
  Effect_Size = c("d = 0.20", "d = 0.35", "d = 0.50", "d = 0.80"),
  Example = c("3 mmHg BP reduction",
              "5 mmHg BP reduction",
              "7.5 mmHg BP reduction",
              "12 mmHg BP reduction"),
  Power_N64 = c("20%", "50%", "80%", "99%"),
  Interpretation = c("Underpowered for small effects",
                     "Moderate power for small-to-medium effects",
                     "Well-powered for medium effects (target)",
                     "Overpowered for large effects"),
  stringsAsFactors = FALSE
)

library(knitr)
kable(interpretation, caption = "Power at Different Effect Sizes (N = 64 per group, α = 0.05)")
```

**Key insights**:

- **Study is well-powered (≥ 80%)** for effects d ≥ 0.50
- **Study is underpowered (< 80%)** for effects d < 0.50
- **Minimal Detectable Effect**: d = 0.50 (medium effect)
- **Implication**: If the true effect is small (d = 0.20), expect low power (20% chance of detecting it)

### Two-Group Comparison (Binary Outcome)

**Scenario**: Comparing two proportions (e.g., treatment success rates).

**Power curve (Power vs Risk Difference)**:

```{r binary-outcome, echo=FALSE}
binary_interp <- data.frame(
  Risk_Difference = c("5%", "10%", "15%", "20%"),
  Example = c("20% vs 25%", "20% vs 30%", "20% vs 35%", "20% vs 40%"),
  Relative_Risk = c("1.25", "1.50", "1.75", "2.00"),
  Power_N200 = c("25%", "55%", "80%", "95%"),
  Interpretation = c("Underpowered for small differences",
                     "Moderate power for moderate differences",
                     "Well-powered for large differences (target)",
                     "Overpowered for very large differences"),
  stringsAsFactors = FALSE
)

kable(binary_interp, caption = "Power at Different Risk Differences (N = 100 per group, α = 0.05)")
```

**Key insights**:

- MDE = 15 percentage points (RR = 1.75)
- Study cannot reliably detect differences < 15 percentage points
- If clinical threshold is 10 percentage points, study is underpowered

### Survival Analysis (Hazard Ratio)

**Scenario**: Time-to-event outcome (e.g., mortality, disease recurrence).

**Power curve (Power vs Hazard Ratio)**:

```{r survival-interpretation, echo=FALSE}
survival_interp <- data.frame(
  Hazard_Ratio = c("1.30", "1.50", "1.75", "2.00"),
  Interpretation = c("30% increased hazard",
                     "50% increased hazard",
                     "75% increased hazard",
                     "100% increased hazard (doubling)"),
  Power_180_Events = c("40%", "70%", "90%", "98%"),
  Clinical_Example = c("Modest treatment effect",
                       "Moderate treatment effect",
                       "Strong treatment effect",
                       "Very strong treatment effect"),
  stringsAsFactors = FALSE
)

kable(survival_interp, caption = "Power at Different Hazard Ratios (180 events, α = 0.05)")
```

**Key insights**:

- Power depends on **number of events**, not total sample size
- MDE ≈ HR = 1.50 (50% increased hazard)
- To detect smaller effects (HR < 1.50), need more events (longer follow-up or higher-risk cohort)

### Matched Case-Control Study

**Scenario**: Comparing exposure odds between matched pairs.

**Power curve (Power vs Odds Ratio)**:

```{r matched-cc, echo=FALSE}
matched_interp <- data.frame(
  Odds_Ratio = c("1.50", "2.00", "2.50", "3.00"),
  Interpretation = c("50% increased odds",
                     "100% increased odds (doubling)",
                     "150% increased odds",
                     "200% increased odds (tripling)"),
  Power_100_Pairs = c("30%", "60%", "80%", "92%"),
  stringsAsFactors = FALSE
)

kable(matched_interp, caption = "Power at Different Odds Ratios (100 matched pairs, α = 0.05)")
```

**Key insights**:

- Matching increases efficiency by controlling for confounders
- MDE ≈ OR = 2.5
- 100 pairs insufficient for detecting OR < 2.5

## Common Power Curve Patterns

### Pattern 1: Steep Curve (High Sensitivity to Effect Size)

**Characteristics**:
- Power increases rapidly with small changes in effect size
- Narrow range between underpowered and well-powered

**Occurs when**:
- Large sample size relative to expected effect
- High event rate (for binary outcomes)
- Many events (for survival analysis)

**Interpretation**: Study is **robust to misspecification** of effect size. Even if true effect is slightly smaller than expected, still adequately powered.

**Example**:
```
Effect Size:  0.40  →  0.45  →  0.50  →  0.55
Power:         50%  →   70%  →   85%  →   94%
```

**Implication**: Recruiting slightly more participants yields substantial power gains.

### Pattern 2: Flat Curve (Low Sensitivity to Effect Size)

**Characteristics**:
- Power increases slowly with changes in effect size
- Wide range between underpowered and well-powered

**Occurs when**:
- Small sample size relative to expected effect
- Low event rate (for binary outcomes)
- Few events (for survival analysis)

**Interpretation**: Study is **vulnerable to misspecification**. If true effect is smaller than expected, power drops dramatically.

**Example**:
```
Effect Size:  0.40  →  0.45  →  0.50  →  0.55
Power:         25%  →   28%  →   32%  →   36%
```

**Implication**: Study is underpowered across a wide range of effect sizes. Need substantial increase in N to achieve adequate power.

### Pattern 3: Early Plateau (Overpowered Study)

**Characteristics**:
- Power reaches 80-90% for small-to-moderate effect sizes
- Power approaches 100% for moderate effects

**Occurs when**:
- Very large sample size
- Overly conservative planning

**Interpretation**: Study is **overpowered** for expected effects. Could reduce N without sacrificing power.

**Example**:
```
Effect Size:  0.20  →  0.30  →  0.40  →  0.50
Power:         75%  →   95%  →   99%  →  >99%
```

**Implication**: Consider reducing sample size to save costs, or reallocate resources to detect smaller effects or add secondary outcomes.

### Pattern 4: Late Plateau (Underpowered Study)

**Characteristics**:
- Power remains low (< 50%) for small-to-moderate effect sizes
- Only reaches 80% for large effects

**Occurs when**:
- Small sample size
- Unrealistic effect size assumptions

**Interpretation**: Study is **underpowered** for realistic effects. Only detect very large effects.

**Example**:
```
Effect Size:  0.20  →  0.40  →  0.60  →  0.80
Power:         10%  →   25%  →   50%  →   75%
```

**Implication**: Increase N substantially or do not proceed (study has low chance of success).

## Using Power Curves for Sensitivity Analysis

### Scenario: Uncertain Effect Size

**Problem**: Literature reports conflicting effect sizes (d = 0.40 to d = 0.60). What sample size ensures adequate power across this range?

**Solution**: Plot power curves for different N and identify which provides ≥ 80% power for d ≥ 0.40.

```{r sensitivity-n, eval=FALSE}
library(pwr)
library(ggplot2)

# Generate power curves for N = 50, 75, 100, 125 per group
effect_sizes <- seq(0.2, 0.8, by = 0.01)
sample_sizes <- c(50, 75, 100, 125)

power_data <- data.frame()
for (n in sample_sizes) {
  for (d in effect_sizes) {
    result <- pwr.t.test(n = n, d = d, sig.level = 0.05, type = "two.sample", alternative = "two.sided")
    power_data <- rbind(power_data, data.frame(N = n, Effect_Size = d, Power = result$power))
  }
}

ggplot(power_data, aes(x = Effect_Size, y = Power, color = factor(N), group = N)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0.40, linetype = "dashed", color = "darkgreen") +
  labs(
    title = "Sensitivity Analysis: Power Across Sample Sizes",
    x = "Effect Size (Cohen's d)",
    y = "Statistical Power",
    color = "N per group"
  ) +
  theme_minimal()
```

**Interpretation**:
- **N = 50**: Power = 52% at d = 0.40 (underpowered)
- **N = 75**: Power = 72% at d = 0.40 (marginal)
- **N = 100**: Power = 85% at d = 0.40 (adequate) ✅
- **N = 125**: Power = 92% at d = 0.40 (robust)

**Decision**: Recruit **N = 100 per group** to ensure ≥ 80% power for effects d ≥ 0.40.

### Scenario: Budget Constraints

**Problem**: Budget allows N = 200 total (100 per group). What effects can we detect?

**Solution**: Plot power curve for N = 100 and identify MDE (effect size where power = 0.80).

```{r budget-constraint, eval=FALSE}
n_fixed <- 100
effect_sizes <- seq(0.2, 1.0, by = 0.01)

power_values <- sapply(effect_sizes, function(d) {
  result <- pwr.t.test(n = n_fixed, d = d, sig.level = 0.05, type = "two.sample", alternative = "two.sided")
  return(result$power)
})

df <- data.frame(Effect_Size = effect_sizes, Power = power_values)

# Find MDE (effect size at 80% power)
mde <- effect_sizes[which.min(abs(power_values - 0.80))]

ggplot(df, aes(x = Effect_Size, y = Power)) +
  geom_line(linewidth = 1.2, color = "#1f77b4") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  geom_vline(xintercept = mde, linetype = "dashed", color = "darkgreen") +
  annotate("text", x = mde + 0.05, y = 0.15, label = paste0("MDE = ", round(mde, 2)), color = "darkgreen") +
  labs(
    title = "Power Curve for Fixed Budget (N = 100 per group)",
    x = "Effect Size (Cohen's d)",
    y = "Statistical Power"
  ) +
  theme_minimal()
```

**Interpretation**:
- MDE ≈ d = 0.40
- Study is well-powered for effects d ≥ 0.40
- Study is underpowered for effects d < 0.40

**Decision**: Compare MDE (0.40) to clinically meaningful threshold. If threshold ≤ 0.40, proceed. If threshold < 0.40, seek additional funding or do not proceed.

### Scenario: Multiple Outcomes

**Problem**: Primary outcome (continuous) and secondary outcome (binary) have different power profiles.

**Solution**: Plot power curves for both outcomes to ensure both are adequately powered.

```{r multiple-outcomes, eval=FALSE}
# Primary: Continuous outcome (d = 0.50 target)
primary_power <- sapply(seq(0.2, 1.0, by = 0.01), function(d) {
  pwr.t.test(n = 80, d = d, sig.level = 0.05, type = "two.sample", alternative = "two.sided")$power
})

# Secondary: Binary outcome (p1 = 0.30, p2 = 0.20, target RD = 0.10)
# Convert to Cohen's h for power calculation
p1_vals <- seq(0.22, 0.40, by = 0.001)
p2 <- 0.20
h_vals <- 2 * (asin(sqrt(p1_vals)) - asin(sqrt(p2)))
secondary_power <- sapply(h_vals, function(h) {
  pwr.2p.test(n = 80, h = h, sig.level = 0.05, alternative = "two.sided")$power
})

# Plot both curves
df_primary <- data.frame(Effect_Size = seq(0.2, 1.0, by = 0.01), Power = primary_power, Outcome = "Primary (Continuous)")
df_secondary <- data.frame(Effect_Size = p1_vals - p2, Power = secondary_power, Outcome = "Secondary (Binary)")

df_combined <- rbind(df_primary, df_secondary)

ggplot(df_combined, aes(x = Effect_Size, y = Power, color = Outcome)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  labs(
    title = "Power Curves: Primary vs Secondary Outcomes (N = 80 per group)",
    x = "Effect Size (Primary: Cohen's d; Secondary: Risk Difference)",
    y = "Statistical Power",
    color = "Outcome"
  ) +
  theme_minimal()
```

**Interpretation**:
- Primary outcome: MDE = d = 0.45 (adequate)
- Secondary outcome: MDE = RD = 0.12 (may be underpowered if clinical threshold is RD = 0.10)

**Decision**: If secondary outcome is critical, increase N to ensure both outcomes are adequately powered. If secondary is exploratory, proceed with N = 80.

## Practical Case Studies with Power Curves

### Case Study 1: Behavioral Intervention Trial

**Scenario**:

A behavioral intervention aims to increase physical activity minutes per week. Literature suggests effect sizes ranging from d = 0.30 (small) to d = 0.60 (medium).

**Research question**: What sample size ensures ≥ 80% power across this range?

**Analysis**:

```{r case1-plot, eval=FALSE}
# Plot power curves for N = 60, 90, 120, 150 per group
sample_sizes <- c(60, 90, 120, 150)
effect_sizes <- seq(0.2, 0.8, by = 0.01)

power_data <- expand.grid(N = sample_sizes, Effect_Size = effect_sizes)
power_data$Power <- mapply(function(n, d) {
  pwr.t.test(n = n, d = d, sig.level = 0.05, type = "two.sample", alternative = "two.sided")$power
}, power_data$N, power_data$Effect_Size)

ggplot(power_data, aes(x = Effect_Size, y = Power, color = factor(N))) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 0.30, linetype = "dashed", color = "darkgreen", alpha = 0.5) +
  geom_vline(xintercept = 0.60, linetype = "dashed", color = "darkgreen", alpha = 0.5) +
  annotate("rect", xmin = 0.30, xmax = 0.60, ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "green") +
  annotate("text", x = 0.45, y = 0.95, label = "Target Range", color = "darkgreen") +
  labs(
    title = "Sensitivity Analysis: Physical Activity Intervention",
    x = "Effect Size (Cohen's d)",
    y = "Statistical Power",
    color = "N per group"
  ) +
  theme_minimal()
```

**Results**:

| N per group | Power at d = 0.30 | Power at d = 0.60 |
|-------------|-------------------|-------------------|
| 60          | 42%               | 97%               |
| 90          | 58%               | 99%               |
| 120         | 71%               | >99%              |
| 150         | 81%               | >99%              |

**Interpretation**:

- **N = 60**: Underpowered for d = 0.30 (42%)
- **N = 90**: Marginal for d = 0.30 (58%)
- **N = 120**: Adequate for d ≥ 0.35 (70%+)
- **N = 150**: Adequate for d ≥ 0.30 (81%) ✅

**Decision**: Recruit **N = 150 per group** (300 total) to ensure ≥ 80% power for effects d ≥ 0.30, covering the full range of plausible effects.

### Case Study 2: Rare Disease Drug Trial

**Scenario**:

A Phase 2 trial can enroll **N = 50 total** (25 per group) due to small eligible population. Outcome is disease severity score (continuous, SD = 10).

**Research question**: What effects can this study detect?

**Analysis**:

```{r case2-plot, eval=FALSE}
n_fixed <- 25
effect_sizes <- seq(0.2, 1.5, by = 0.01)

power_values <- sapply(effect_sizes, function(d) {
  pwr.t.test(n = n_fixed, d = d, sig.level = 0.05, type = "two.sample", alternative = "two.sided")$power
})

df <- data.frame(Effect_Size = effect_sizes, Power = power_values)

# Find MDE
mde <- effect_sizes[which.min(abs(power_values - 0.80))]

# Clinical threshold: 5-point reduction = 0.5 SD = d = 0.50
clinical_threshold <- 0.50

ggplot(df, aes(x = Effect_Size, y = Power)) +
  geom_line(linewidth = 1.2, color = "#1f77b4") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  geom_vline(xintercept = mde, linetype = "dashed", color = "purple") +
  geom_vline(xintercept = clinical_threshold, linetype = "dashed", color = "darkgreen") +
  annotate("text", x = mde + 0.1, y = 0.85, label = paste0("MDE = ", round(mde, 2)), color = "purple") +
  annotate("text", x = clinical_threshold + 0.1, y = 0.30, label = "Clinical Threshold\n(d = 0.50)", color = "darkgreen") +
  annotate("rect", xmin = -Inf, xmax = mde, ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "red") +
  annotate("text", x = 0.50, y = 0.95, label = "Underpowered Region", color = "red") +
  labs(
    title = "Power Curve: Rare Disease Trial (N = 25 per group)",
    x = "Effect Size (Cohen's d)",
    y = "Statistical Power"
  ) +
  theme_minimal()
```

**Results**:

- MDE ≈ **d = 0.81** (large effect)
- Clinical threshold: d = 0.50 (medium effect)
- **Gap**: MDE > clinical threshold (underpowered)

**Interpretation**:

- Study can only detect **large effects (d ≥ 0.81)**
- **Cannot detect clinically meaningful medium effects (d = 0.50)**
- At clinical threshold (d = 0.50), power ≈ **40%**

**Decision options**:

1. **Multi-site collaboration**: Increase to N = 64 per group to achieve MDE = d = 0.50
2. **Proceed as pilot**: Accept underpowering, label as exploratory Phase 2
3. **Do not proceed**: Insufficient power for meaningful inference

**Final decision**: Proceed as **exploratory pilot** with N = 50, using Bayesian analysis to incorporate prior information. Plan larger Phase 3 trial if results promising.

### Case Study 3: Cluster-Randomized Trial

**Scenario**:

Quality improvement intervention across hospitals. Due to clustering (ICC = 0.05, m = 50 per hospital), effective sample size is reduced by design effect (DE = 3.45).

**Research question**: How does clustering affect power curve?

**Analysis**:

```{r case3-plot, eval=FALSE}
# Scenario A: Individual randomization (no clustering)
n_individual <- 100  # per group
effect_sizes <- seq(0.2, 0.8, by = 0.01)
power_individual <- sapply(effect_sizes, function(d) {
  pwr.t.test(n = n_individual, d = d, sig.level = 0.05, type = "two.sample", alternative = "two.sided")$power
})

# Scenario B: Cluster randomization
# Design effect = 3.45, so effective N = 100 / 3.45 ≈ 29
n_effective <- round(n_individual / 3.45)
power_clustered <- sapply(effect_sizes, function(d) {
  pwr.t.test(n = n_effective, d = d, sig.level = 0.05, type = "two.sample", alternative = "two.sided")$power
})

df <- data.frame(
  Effect_Size = rep(effect_sizes, 2),
  Power = c(power_individual, power_clustered),
  Design = rep(c("Individual Randomization", "Cluster Randomization (DE=3.45)"), each = length(effect_sizes))
)

ggplot(df, aes(x = Effect_Size, y = Power, color = Design)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  labs(
    title = "Impact of Clustering on Power Curve (N = 100 per group)",
    x = "Effect Size (Cohen's d)",
    y = "Statistical Power",
    color = "Design Type"
  ) +
  theme_minimal()
```

**Results**:

| Effect Size | Individual Power | Clustered Power | Power Loss |
|-------------|------------------|-----------------|------------|
| d = 0.40    | 85%              | 34%             | -51 points |
| d = 0.50    | 97%              | 48%             | -49 points |
| d = 0.60    | 99%              | 62%             | -37 points |
| d = 0.80    | >99%             | 85%             | -14 points |

**Interpretation**:

- Clustering **dramatically reduces power** (by 30-50 percentage points for moderate effects)
- MDE increases from d = 0.40 (individual) to d = 0.78 (clustered)
- To maintain 80% power for d = 0.50, need to inflate N by design effect: 100 × 3.45 = **345 per group**

**Decision**: Account for design effect during planning. If recruiting 345 per group infeasible, consider:
- Reducing cluster size (m = 50 → 25) to reduce design effect (3.45 → 2.20)
- Using individual randomization if contamination not a concern

## Advanced Visualizations

### Multi-Dimensional Power Curves

For studies with **multiple adjustable parameters** (sample size, effect size, alpha, allocation ratio), create **contour plots** or **3D surfaces**.

**Example: Power as a function of N and effect size**:

```{r contour-plot, eval=FALSE}
library(ggplot2)

# Generate grid
n_values <- seq(30, 150, by = 5)
d_values <- seq(0.2, 0.8, by = 0.02)
power_grid <- expand.grid(N = n_values, Effect_Size = d_values)

power_grid$Power <- mapply(function(n, d) {
  pwr.t.test(n = n, d = d, sig.level = 0.05, type = "two.sample", alternative = "two.sided")$power
}, power_grid$N, power_grid$Effect_Size)

# Contour plot
ggplot(power_grid, aes(x = N, y = Effect_Size, z = Power)) +
  geom_contour_filled(breaks = seq(0, 1, by = 0.1)) +
  geom_contour(color = "white", alpha = 0.3, breaks = c(0.80)) +  # Highlight 80% power
  labs(
    title = "Power Contour Plot: Sample Size vs Effect Size",
    x = "Sample Size (N per group)",
    y = "Effect Size (Cohen's d)",
    fill = "Power"
  ) +
  theme_minimal()
```

**Interpretation**: Contour lines show **combinations of N and d that yield the same power**. The 80% power contour is the **boundary between underpowered and adequately powered regions**.

### Interactive Power Curves

The Power Analysis Tool provides **interactive plotly visualizations** that allow:

- **Hovering** to see exact power values at any point
- **Zooming** to focus on regions of interest
- **Toggling** between power vs effect size and power vs sample size views
- **Comparing** multiple scenarios simultaneously (e.g., with/without missing data adjustment)

**Benefits**:
- Explore trade-offs dynamically
- Communicate results to stakeholders visually
- Export high-resolution plots for publications

## Summary

**Key takeaways**:

1. **Power curves are more informative** than single-point calculations — show power across range of effect sizes or sample sizes

2. **Anatomy of power curves**:
   - Sigmoid shape (slow → rapid → plateau)
   - Crosses 80% at Minimal Detectable Effect (MDE)
   - Alpha level at null (effect size = 0)

3. **Common patterns**:
   - **Steep curve**: Robust to effect size uncertainty
   - **Flat curve**: Vulnerable to misspecification
   - **Early plateau**: Overpowered (can reduce N)
   - **Late plateau**: Underpowered (increase N or abandon)

4. **Use power curves for**:
   - Sensitivity analysis (effect size uncertainty)
   - Budget constraints (fixed N, determine MDE)
   - Comparing designs (clustering, missing data, allocation ratios)
   - Communication with stakeholders

5. **Practical workflow**:
   - Plot power curve for proposed N
   - Identify MDE (where curve crosses 80%)
   - Compare MDE to clinical threshold
   - Adjust N if needed to ensure MDE ≤ threshold

6. **Interactive visualizations** (in Power Analysis Tool) facilitate exploration and decision-making

## References

1. Cohen J (1988). *Statistical Power Analysis for the Behavioral Sciences* (2nd ed.). Routledge.

2. Lenth RV (2001). Some practical guidelines for effective sample size determination. *The American Statistician* 55(3):187-193.

3. Button KS, Ioannidis JPA, Mokrysz C, et al. (2013). Power failure: why small sample size undermines the reliability of neuroscience. *Nature Reviews Neuroscience* 14:365-376.

4. Gelman A, Carlin J (2014). Beyond power calculations: Assessing Type S (sign) and Type M (magnitude) errors. *Perspectives on Psychological Science* 9(6):641-651.

5. Dziak JJ, Dierker LC, Abar B (2020). The interpretation of statistical power after the data have been gathered. *Current Psychology* 39:870-877.

## Further Reading

- See `vignette("minimal-detectable-effects")` for understanding feasibility constraints
- See `vignette("design-effects-clustering")` for impact of clustering on power curves
- See `vignette("missing-data-adjustments")` for impact of missing data on power curves
- Explore all tabs in the Power Analysis Tool for interactive power curve visualizations
