---
title: "Minimal Detectable Effect Sizes: Understanding Feasibility Constraints"
author: "Power Analysis Tool Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Minimal Detectable Effect Sizes: Understanding Feasibility Constraints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Traditional sample size calculations answer the question: *"How many participants do I need to detect an effect of size X?"*

But in real-world research, sample size is often **constrained** by:

- **Budget limitations**: Limited funding for recruitment and follow-up
- **Feasibility**: Small eligible patient population
- **Time constraints**: Short study duration limits enrollment
- **Logistical barriers**: Limited sites, staff, or resources

In these scenarios, a more practical question is: *"Given my constrained sample size, what is the smallest effect I can reliably detect?"*

This is the **Minimal Detectable Effect Size (MDE)** — also called the **Minimum Detectable Effect**.

This vignette explains:

1. **What is a minimal detectable effect** and why it matters
2. **How to calculate MDE** for different study designs
3. **How to interpret MDE** in the context of clinical or practical significance
4. **When to use MDE** vs traditional sample size calculations
5. **Practical workflows** and decision-making examples

## What is a Minimal Detectable Effect?

### Definition

The **Minimal Detectable Effect Size (MDE)** is the smallest true effect that a study can reliably detect (with specified power and significance level) given a fixed sample size.

**Key components**:

- **Fixed N**: Sample size is predetermined (constrained by feasibility)
- **Fixed α**: Significance level (typically 0.05)
- **Fixed power**: Desired statistical power (typically 0.80 or 0.90)
- **Unknown effect size**: What we're solving for

### Why MDE Matters

Traditional power analysis assumes:
1. You specify the effect size of interest (from literature, pilot data, or clinical importance)
2. Calculate required sample size
3. Recruit that many participants

**Problem**: What if the calculated N is 5,000 but you can only afford to recruit 500?

**Solution**: Calculate the MDE for N = 500 to understand:
- What effect sizes are you powered to detect?
- Is the MDE clinically meaningful?
- Should you proceed with the study or seek additional resources/collaborations?

### MDE as a Reality Check

The MDE provides a **feasibility reality check**:

**Scenario A: Feasible Study**
- Can recruit N = 400
- MDE = 10 percentage points difference (e.g., 30% vs 40%)
- Clinically meaningful threshold = 8 percentage points
- **Decision**: Proceed — study is powered to detect meaningful effects

**Scenario B: Underpowered Study**
- Can recruit N = 100
- MDE = 25 percentage points (e.g., 30% vs 55%)
- Clinically meaningful threshold = 8 percentage points
- **Decision**: Do NOT proceed — study can only detect unrealistically large effects

## Calculating MDE for Different Designs

### Single Proportion

**Question**: Given fixed N, what is the smallest difference from a null proportion we can detect?

**Example**:
- Historical proportion: **p₀ = 0.30** (30% event rate)
- Available sample: **N = 150**
- Power: **80%**, α = 0.05

**Calculation (using Power Analysis Tool)**:

```{r mde-single-prop, eval=FALSE}
# This calculation is performed via binary search in the tool
# For a single proportion test:

library(pwr)

# Solve for effect size h given fixed n
result <- pwr.p.test(
  n = 150,
  sig.level = 0.05,
  power = 0.80,
  alternative = "two.sided"
)

h_mde <- result$h  # Cohen's h

# Convert h back to proportion
# h = 2 * (arcsin(sqrt(p1)) - arcsin(sqrt(p0)))
# For p0 = 0.30, solve for p1 using h_mde
p0 <- 0.30
p1_mde <- (sin(asin(sqrt(p0)) + h_mde / 2))^2

cat("With N =", 150, ", minimal detectable proportion is p1 =", round(p1_mde, 3), "\n")
cat("Minimal detectable difference:", round(p1_mde - p0, 3), "\n")
```

**Typical output**:
```
With N = 150, minimal detectable proportion is p1 = 0.436
Minimal detectable difference: 0.136 (13.6 percentage points)
```

**Interpretation**: With 150 participants, the study can detect an increase from 30% to at least 44% (13.6 percentage point difference). Smaller effects will not be reliably detected.

### Two-Group Comparison (Proportions)

**Question**: Given fixed N per group, what is the smallest difference between two proportions we can detect?

**Example**:
- Control group proportion: **p₂ = 0.20** (20% event rate)
- Available sample: **N = 200** (100 per group)
- Power: **80%**, α = 0.05

**Calculation**:

```{r mde-two-prop, eval=FALSE}
result <- pwr.2p.test(
  n = 100,  # Per group
  sig.level = 0.05,
  power = 0.80,
  alternative = "two.sided"
)

h_mde <- result$h

# Convert h to proportions (p2 = 0.20 is fixed)
p2 <- 0.20
# h = 2 * (arcsin(sqrt(p1)) - arcsin(sqrt(p2)))
# Solve for p1
p1_mde <- (sin(asin(sqrt(p2)) + h_mde / 2))^2

cat("With N = 100 per group, minimal detectable p1 =", round(p1_mde, 3), "\n")
cat("Minimal detectable difference:", round(p1_mde - p2, 3), "\n")

# Calculate effect measures
rr_mde <- p1_mde / p2
or_mde <- (p1_mde / (1 - p1_mde)) / (p2 / (1 - p2))
rd_mde <- p1_mde - p2

cat("MDE in terms of:\n")
cat("  Risk Difference (RD):", round(rd_mde, 3), "\n")
cat("  Relative Risk (RR):", round(rr_mde, 2), "\n")
cat("  Odds Ratio (OR):", round(or_mde, 2), "\n")
```

**Typical output**:
```
With N = 100 per group, minimal detectable p1 = 0.361
Minimal detectable difference: 0.161 (16.1 percentage points)

MDE in terms of:
  Risk Difference (RD): 0.161
  Relative Risk (RR): 1.81
  Odds Ratio (OR): 2.27
```

**Interpretation**: With 100 per group, the study can detect an increase from 20% to at least 36% (RR ≥ 1.81, OR ≥ 2.27). Smaller effects will not be reliably detected.

### Continuous Outcomes (t-test)

**Question**: Given fixed N per group, what is the smallest mean difference (in standardized units) we can detect?

**Example**:
- Available sample: **N = 120** (60 per group)
- Power: **80%**, α = 0.05

**Calculation**:

```{r mde-continuous, eval=FALSE}
result <- pwr.t.test(
  n = 60,  # Per group
  sig.level = 0.05,
  power = 0.80,
  type = "two.sample",
  alternative = "two.sided"
)

d_mde <- result$d

cat("With N = 60 per group, minimal detectable Cohen's d =", round(d_mde, 3), "\n")

# Interpretation in terms of raw differences (if SD is known)
# Example: If outcome is blood pressure with SD = 15 mmHg
sd <- 15
mean_diff_mde <- d_mde * sd

cat("If SD = 15 mmHg, minimal detectable mean difference:", round(mean_diff_mde, 1), "mmHg\n")
```

**Typical output**:
```
With N = 60 per group, minimal detectable Cohen's d = 0.519

If SD = 15 mmHg, minimal detectable mean difference: 7.8 mmHg
```

**Interpretation**: With 60 per group, the study can detect a standardized mean difference of at least 0.52 (a "medium" effect by Cohen's conventions). In practical terms, this corresponds to a 7.8 mmHg blood pressure difference if SD = 15 mmHg.

### Survival Analysis (Cox Regression)

**Question**: Given fixed sample size and event rate, what is the smallest hazard ratio we can detect?

**Example**:
- Available sample: **N = 300**
- Expected event rate: **60%** (180 events)
- Power: **80%**, α = 0.05

**Calculation (conceptual)**:

For Cox regression, power depends on the **number of events**, not total sample size.

```{r mde-survival, eval=FALSE}
# Approximate calculation using event-driven power
# For a two-group comparison with equal allocation:

n_events <- 180
z_alpha <- qnorm(1 - 0.05 / 2)  # 1.96
z_beta <- qnorm(0.80)           # 0.84

# Formula: log(HR) ≈ (z_alpha + z_beta) / sqrt(n_events / 4)
log_hr_mde <- (z_alpha + z_beta) / sqrt(n_events / 4)
hr_mde <- exp(log_hr_mde)

cat("With", n_events, "events, minimal detectable HR =", round(hr_mde, 2), "\n")
```

**Typical output**:
```
With 180 events, minimal detectable HR = 1.53
```

**Interpretation**: With 300 participants and 180 events, the study can detect a hazard ratio of at least 1.53 (53% increased hazard). Smaller hazard ratios will not be reliably detected.

### Matched Case-Control Studies

**Question**: Given fixed number of matched pairs, what is the smallest odds ratio we can detect?

**Example**:
- Available sample: **100 matched pairs**
- Exposure proportion in controls: **30%**
- Matching ratio: **1:1**
- Power: **80%**, α = 0.05

**Calculation (using Power Analysis Tool)**:

The tool uses binary search to find the minimal OR that achieves 80% power given fixed N. This involves iteratively testing OR values until the power curve crosses 0.80.

**Typical MDE**: OR ≥ 2.0 to 3.0 for 100 matched pairs (depends on baseline exposure).

### Non-Inferiority Trials

**Question**: Given fixed sample size, what is the widest non-inferiority margin we can demonstrate?

**Example**:
- Expected event rates: **p₁ = 0.15** (test), **p₂ = 0.15** (reference)
- Available sample: **N = 400** (200 per group)
- Power: **80%**, α = 0.025 (one-sided)

**Calculation**:

The tool uses binary search to find the maximum non-inferiority margin (Δ) such that the upper bound of the confidence interval for (p₁ - p₂) excludes Δ with 80% power.

**Typical MDE**: Margin ≤ 5-10 percentage points for N = 400.

## Interpreting MDE in Context

### Clinical vs Statistical Significance

The MDE must be interpreted relative to **clinical or practical significance thresholds**:

**Example 1: Blood Pressure Reduction**

- MDE: 7.8 mmHg reduction
- Clinically meaningful: 5 mmHg reduction (established guideline threshold)
- **Interpretation**: Study is adequately powered; MDE is smaller than clinical threshold ✅

**Example 2: Diabetes HbA1c Reduction**

- MDE: 1.2% reduction
- Clinically meaningful: 0.5% reduction (established threshold)
- **Interpretation**: Study is underpowered; MDE exceeds clinical threshold ❌

### Effect Size Benchmarks

**Cohen's d (Continuous Outcomes)**:

```{r cohens-d-table, echo=FALSE}
cohens_table <- data.frame(
  Effect_Size = c("Small", "Medium", "Large"),
  Cohens_d = c("0.20", "0.50", "0.80"),
  Example = c("Small but noticeable (e.g., 3 mmHg BP)",
              "Moderate, clinically relevant (e.g., 7.5 mmHg BP)",
              "Large, clinically important (e.g., 12 mmHg BP)"),
  stringsAsFactors = FALSE
)

library(knitr)
kable(cohens_table, caption = "Cohen's d Benchmarks for Continuous Outcomes")
```

**Relative Risk / Odds Ratio (Binary Outcomes)**:

```{r rr-or-table, echo=FALSE}
effect_table <- data.frame(
  Effect_Size = c("Small", "Moderate", "Large", "Very Large"),
  Relative_Risk = c("1.1 - 1.3", "1.3 - 2.0", "2.0 - 3.0", "> 3.0"),
  Odds_Ratio = c("1.2 - 1.5", "1.5 - 2.5", "2.5 - 4.0", "> 4.0"),
  Example = c("Weak association (e.g., minor risk factor)",
              "Moderate association (e.g., typical treatment effect)",
              "Strong association (e.g., effective prevention)",
              "Very strong (e.g., smoking → lung cancer)"),
  stringsAsFactors = FALSE
)

kable(effect_table, caption = "Relative Effect Size Benchmarks for Binary Outcomes")
```

**Hazard Ratio (Survival Analysis)**:

Similar interpretation to RR/OR:

- **HR = 1.0**: No effect
- **HR = 1.3**: 30% increased hazard (moderate)
- **HR = 2.0**: 100% increased hazard (large)

### Domain-Specific Thresholds

Different fields have established thresholds for meaningful effects:

**Cardiovascular Medicine**:
- Blood pressure: ≥ 5 mmHg reduction
- LDL cholesterol: ≥ 10% reduction
- Cardiovascular events: RR ≤ 0.90 (10% reduction)

**Diabetes**:
- HbA1c: ≥ 0.5% reduction
- Fasting glucose: ≥ 10 mg/dL reduction

**Mental Health**:
- Depression scales (e.g., PHQ-9): ≥ 3-point change
- Anxiety scales (e.g., GAD-7): ≥ 4-point change

**Education**:
- Standardized test scores: ≥ 0.20 SD improvement

**Recommendation**: Always specify **a priori** what effect size is clinically/practically meaningful, then compare MDE to this threshold.

## Decision-Making Framework

### Step 1: Calculate MDE for Your Constrained N

Use the Power Analysis Tool or statistical software to determine the MDE given:

- Fixed sample size (N)
- Desired power (typically 0.80)
- Significance level (α = 0.05)

```{r step1-example, eval=FALSE}
# Example: Two-group comparison with constrained N = 200 (100 per group)
library(pwr)

result <- pwr.2p.test(
  n = 100,
  sig.level = 0.05,
  power = 0.80,
  alternative = "two.sided"
)

h_mde <- result$h
# Convert to practical metrics (RD, RR, OR) based on your baseline p2
```

### Step 2: Compare MDE to Clinical Threshold

**If MDE ≤ Clinical Threshold**: Study is feasible ✅

- The smallest effect you can detect is smaller than (or equal to) what's clinically meaningful
- Proceed with the study

**If MDE > Clinical Threshold**: Study is underpowered ❌

- You can only detect effects larger than what's clinically meaningful
- Risk of **false negatives**: Missing true but smaller effects
- Consider alternatives (Step 3)

### Step 3: Explore Alternatives if Underpowered

**Option A: Increase Sample Size**

- Seek additional funding
- Extend recruitment period
- Add recruitment sites
- Use multi-site collaborations

```{r option-a, eval=FALSE}
# Calculate required N to detect clinical threshold effect
# Example: Need to detect RD = 0.10 (10 percentage points)

# Convert RD to Cohen's h
p1 <- 0.30  # Clinical threshold
p2 <- 0.20
h_target <- 2 * (asin(sqrt(p1)) - asin(sqrt(p2)))

# Calculate required N
result_needed <- pwr.2p.test(
  h = h_target,
  sig.level = 0.05,
  power = 0.80,
  alternative = "two.sided"
)

n_per_group_needed <- ceiling(result_needed$n)
cat("To detect RD = 0.10, need", n_per_group_needed, "per group\n")
```

**Option B: Increase Power for Larger (but Still Meaningful) Effects**

If you can't detect the smallest meaningful effect, determine what larger effect you CAN detect and assess if it's still meaningful.

```{r option-b, eval=FALSE}
# Calculate MDE with current N but for higher power (e.g., 90%)
result_high_power <- pwr.2p.test(
  n = 100,
  sig.level = 0.05,
  power = 0.90,  # Higher power
  alternative = "two.sided"
)

h_mde_90 <- result_high_power$h
# Compare: with 90% power, MDE is slightly larger
```

**Option C: Accept Lower Power**

If the MDE is close to the clinical threshold, you might accept lower power (e.g., 70%) to make the study feasible.

**Caution**: Lower power increases the risk of false negatives (Type II errors).

```{r option-c, eval=FALSE}
# Calculate MDE with lower power (70%)
result_low_power <- pwr.2p.test(
  n = 100,
  sig.level = 0.05,
  power = 0.70,
  alternative = "two.sided"
)

h_mde_70 <- result_low_power$h
# With lower power, MDE is smaller (but risk of missing true effects increases)
```

**Option D: Combine with Meta-Analysis**

If your study alone is underpowered, plan to contribute to a future meta-analysis with other similar studies.

**Option E: Revise Research Question**

If the primary outcome is not feasible, consider:

- **Composite outcomes**: Combine multiple related outcomes to increase event rate
- **Alternative outcomes**: Use a more sensitive outcome (e.g., continuous biomarker instead of binary clinical event)
- **Subgroup analysis**: Focus on higher-risk subgroup with higher event rate

**Option F: Do Not Proceed**

If none of the above are feasible and the MDE is much larger than meaningful, it may be more ethical and efficient to **not conduct the study**.

**Why?**
- Wastes resources (money, time, participant burden)
- Contributes to research waste (underpowered studies clutter literature)
- Ethical concerns (exposing participants to interventions without adequate chance of detecting benefit)

## Practical Case Studies

### Case Study 1: Community Health Intervention

**Scenario**:

A community health center wants to test a smoking cessation intervention.

**Parameters**:
- Outcome: Smoking cessation at 6 months (binary)
- Historical quit rate: **15%**
- Available budget: **200 participants** (100 per group)
- Clinically meaningful: **10 percentage point increase** (to 25%)

**Analysis**:

```{r case1-analysis, eval=FALSE}
# Calculate MDE with N = 100 per group
result <- pwr.2p.test(
  n = 100,
  sig.level = 0.05,
  power = 0.80,
  alternative = "two.sided"
)

h_mde <- result$h
p2 <- 0.15

# Solve for p1 given h_mde and p2
p1_mde <- (sin(asin(sqrt(p2)) + h_mde / 2))^2
rd_mde <- p1_mde - p2

cat("MDE (Risk Difference):", round(rd_mde, 3), "\n")
cat("MDE (Cessation rate):", round(p1_mde, 3), "(", round(p1_mde * 100, 1), "%)\n")
cat("Clinical threshold: 0.10 (10 percentage points)\n")

if (rd_mde <= 0.10) {
  cat("✅ Study is feasible: MDE ≤ clinical threshold\n")
} else {
  cat("❌ Study is underpowered: MDE > clinical threshold\n")
}
```

**Output**:
```
MDE (Risk Difference): 0.159
MDE (Cessation rate): 0.309 (30.9%)
Clinical threshold: 0.10 (10 percentage points)

❌ Study is underpowered: MDE > clinical threshold
```

**Interpretation**:

- With 100 per group, can only detect an increase from 15% to 31% (16 percentage points)
- Clinical threshold is 10 percentage points (15% → 25%)
- **Conclusion**: Study cannot detect the clinically meaningful 10-point increase

**Decision Options**:

1. **Increase N**: Calculate required sample size for 10-point increase
   ```{r case1-option1, eval=FALSE}
   h_target <- 2 * (asin(sqrt(0.25)) - asin(sqrt(0.15)))
   result_target <- pwr.2p.test(h = h_target, sig.level = 0.05, power = 0.80, alternative = "two.sided")
   n_needed <- ceiling(result_target$n)
   cat("Need", n_needed, "per group (", n_needed * 2, "total)\n")
   # Output: Need 293 per group (586 total)
   ```

2. **Accept larger MDE**: Reframe research question as detecting 15% → 31% increase (still meaningful?)

3. **Do not proceed**: If 586 participants infeasible and 16-point increase not meaningful, cancel study

**Final Decision**: Seek additional funding to recruit 600 participants, or do not proceed.

### Case Study 2: Clinical Trial with Fixed Budget

**Scenario**:

A pharmaceutical company has a fixed budget for a Phase 2 trial of a new antihypertensive drug.

**Parameters**:
- Outcome: Change in systolic blood pressure (continuous)
- Baseline SD: **15 mmHg** (from literature)
- Available budget: **N = 120** (60 per group)
- Clinically meaningful: **5 mmHg reduction** (guideline threshold)

**Analysis**:

```{r case2-analysis, eval=FALSE}
# Calculate MDE (Cohen's d) with N = 60 per group
result <- pwr.t.test(
  n = 60,
  sig.level = 0.05,
  power = 0.80,
  type = "two.sample",
  alternative = "two.sided"
)

d_mde <- result$d

# Convert to mmHg
sd <- 15
mean_diff_mde <- d_mde * sd

cat("MDE (Cohen's d):", round(d_mde, 3), "\n")
cat("MDE (mmHg):", round(mean_diff_mde, 2), "mmHg\n")
cat("Clinical threshold: 5 mmHg\n")

if (mean_diff_mde <= 5) {
  cat("✅ Study is feasible: MDE ≤ clinical threshold\n")
} else {
  cat("❌ Study is underpowered: MDE > clinical threshold\n")
}
```

**Output**:
```
MDE (Cohen's d): 0.519
MDE (mmHg): 7.79 mmHg
Clinical threshold: 5 mmHg

❌ Study is underpowered: MDE > clinical threshold
```

**Interpretation**:

- With 60 per group, can only detect a 7.8 mmHg reduction
- Clinical threshold is 5 mmHg
- **Conclusion**: Study cannot detect the clinically meaningful 5 mmHg reduction

**Decision Options**:

1. **Increase N**:
   ```{r case2-option1, eval=FALSE}
   d_target <- 5 / 15  # 0.333
   result_target <- pwr.t.test(d = d_target, sig.level = 0.05, power = 0.80, type = "two.sample", alternative = "two.sided")
   n_needed <- ceiling(result_target$n)
   cat("Need", n_needed, "per group (", n_needed * 2, "total)\n")
   # Output: Need 143 per group (286 total)
   ```

2. **Use a more sensitive outcome**: Switch from office BP to 24-hour ambulatory BP monitoring (lower variance)

3. **Proceed with N = 120 as exploratory study**: Accept that study is powered for 7.8 mmHg, but report results regardless (exploratory Phase 2)

**Final Decision**: Proceed with N = 120 as **exploratory Phase 2** trial. If results show ≥ 7.8 mmHg reduction, conduct larger Phase 3 trial (N = 286) powered for 5 mmHg.

### Case Study 3: Rare Disease Study

**Scenario**:

A rare disease registry can only enroll **50 participants total** (25 per group) for a treatment comparison.

**Parameters**:
- Outcome: Disease severity score (continuous, SD = 10)
- Available sample: **N = 50** (25 per group)
- Clinically meaningful: **5-point reduction** (0.5 SD)

**Analysis**:

```{r case3-analysis, eval=FALSE}
result <- pwr.t.test(
  n = 25,
  sig.level = 0.05,
  power = 0.80,
  type = "two.sample",
  alternative = "two.sided"
)

d_mde <- result$d
sd <- 10
mean_diff_mde <- d_mde * sd

cat("MDE (Cohen's d):", round(d_mde, 3), "\n")
cat("MDE (score points):", round(mean_diff_mde, 2), "\n")
cat("Clinical threshold: 5 points\n")
```

**Output**:
```
MDE (Cohen's d): 0.814
MDE (score points): 8.14
Clinical threshold: 5 points

❌ Study is underpowered: MDE > clinical threshold
```

**Interpretation**:

- With 25 per group, can only detect an 8.1-point reduction
- Clinical threshold is 5 points
- **Conclusion**: Study cannot detect the clinically meaningful 5-point reduction

**Decision Options**:

1. **Multi-site collaboration**: Partner with other registries to increase N to 143 per group (from Case 2 calculation)

2. **Accept high MDE and proceed**: Rare disease trials are inherently underpowered; proceed as pilot/proof-of-concept

3. **Use Bayesian approach**: Use Bayesian methods to incorporate prior information and gain efficiency

**Final Decision**: Proceed with N = 50 as **pilot study**. Use Bayesian analysis incorporating literature priors. Report findings to contribute to future meta-analysis.

## Advanced Topics

### MDE vs Post-Hoc Power

**Important distinction**:

- **MDE (a priori)**: Calculated **before** data collection, using fixed N to determine smallest detectable effect
- **Post-hoc power**: Calculated **after** data collection, using observed effect size to determine power

**Never calculate post-hoc power** for your study's observed effect:

- Circular reasoning: Post-hoc power is a 1:1 function of the p-value
- If p < 0.05, post-hoc power will be high (misleading)
- If p > 0.05, post-hoc power will be low (obvious from non-significance)

**Instead**: Report **MDE** in your study protocol and publications to show what effects the study was designed to detect.

### MDE with Adjustments (Clustering, Missing Data)

When sample size is constrained and you have clustering or missing data, the **effective sample size** is reduced, increasing the MDE.

**Example**:
- Available sample: N = 400
- Clustering: Design effect = 2.0 (ICC = 0.05, m = 20)
- Effective N: 400 / 2.0 = **200**

**MDE calculation**:

```{r mde-adjusted, eval=FALSE}
# Calculate MDE using effective N
n_effective <- 200 / 2  # 100 per group

result <- pwr.2p.test(
  n = n_effective,
  sig.level = 0.05,
  power = 0.80,
  alternative = "two.sided"
)

# MDE will be larger due to reduced effective N
```

**Recommendation**: Always account for clustering and missing data when calculating MDE for constrained samples.

### MDE for Non-Inferiority Trials

Non-inferiority trials aim to show that a new treatment is "not worse" than standard by more than a margin Δ.

**MDE question**: Given constrained N, what is the widest margin we can demonstrate?

**Interpretation**:
- **Narrow margin** (e.g., 2%): Stringent non-inferiority (requires large N)
- **Wide margin** (e.g., 10%): Lenient non-inferiority (feasible with smaller N)

**Decision**: Compare MDE margin to clinically acceptable margin (often half of the superiority threshold).

## Summary

**Key takeaways**:

1. **MDE answers**: "What's the smallest effect I can detect with my constrained sample size?"

2. **Calculate MDE when**:
   - Sample size is fixed by budget, feasibility, or logistics
   - You need a reality check on what effects are detectable

3. **Compare MDE to clinical thresholds**:
   - If MDE ≤ threshold: Study is feasible ✅
   - If MDE > threshold: Study is underpowered ❌

4. **Decision options for underpowered studies**:
   - Increase N (seek funding, extend timeline, add sites)
   - Revise research question (larger effects, alternative outcomes)
   - Accept underpowering (pilot/exploratory study)
   - Do not proceed (avoid research waste)

5. **Always report MDE in study protocols** to show transparency about what effects the study can detect

6. **Never calculate post-hoc power** — use MDE instead

**Practical workflow**:

1. Determine constrained sample size (N)
2. Calculate MDE for N (using Power Analysis Tool or pwr package)
3. Compare MDE to clinically meaningful threshold
4. Decide: proceed, modify, or cancel study
5. Report MDE in protocol and publications

## References

1. Lenth RV (2001). Some practical guidelines for effective sample size determination. *The American Statistician* 55(3):187-193.

2. Gelman A, Carlin J (2014). Beyond power calculations: Assessing Type S (sign) and Type M (magnitude) errors. *Perspectives on Psychological Science* 9(6):641-651.

3. Cohen J (1988). *Statistical Power Analysis for the Behavioral Sciences* (2nd ed.). Routledge.

4. Freedman LS, Lowe D, Macaskill P (1984). Stopping rules for clinical trials incorporating clinical opinion. *Biometrics* 40(3):575-586.

5. Hoenig JM, Heisey DM (2001). The abuse of power: The pervasive fallacy of power calculations for data analysis. *The American Statistician* 55(1):19-24.

## Further Reading

- See `vignette("interactive-power-curves")` for visualizing power across a range of effect sizes
- See `vignette("design-effects-clustering")` for adjusting MDE for clustered data
- See `vignette("missing-data-adjustments")` for adjusting MDE for missing data
- See all tabs in the Power Analysis Tool for interactive MDE calculations
