---
title: "Missing Data Adjustment Strategies for Sample Size Calculations"
author: "Power Analysis Tool Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing Data Adjustment Strategies for Sample Size Calculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

Missing data is ubiquitous in real-world studies. Failing to account for missingness during sample size planning can lead to underpowered studies that cannot detect true effects. This vignette explains:

1. **Why missing data reduces power** and effective sample size
2. **Three missing data mechanisms** (MCAR, MAR, MNAR) and their implications
3. **Two analysis approaches**: Complete Case Analysis (CCA) vs Multiple Imputation (MI)
4. **How to calculate sample size inflation** for each approach
5. **Practical decision-making** for choosing between CCA and MI

## Why Missing Data Matters for Power

Statistical power depends on **effective sample size** — the number of observations actually used in analysis. Missing data reduces this:

- **Lost observations**: Each missing case reduces the dataset
- **Reduced precision**: Smaller sample → wider confidence intervals → lower power
- **Increased variance**: Uncertainty increases with less complete data

**Key principle**: Always plan for more participants than theoretically required to account for expected missingness.

## Missing Data Mechanisms

Understanding the **mechanism** of missingness is critical for choosing the right approach and interpreting results.

### MCAR: Missing Completely at Random

**Definition**: Missingness is unrelated to any observed or unobserved variables.

**Example**: A lab machine randomly malfunctions, causing 10% of blood samples to be lost.

**Properties**:
- Complete cases are a random sample of the full data
- No bias introduced (only loss of power)
- CCA is valid but inefficient
- MI offers efficiency gains

**Analogy**: Randomly flipping a coin to decide whether to measure someone — the flip has nothing to do with their characteristics or outcomes.

### MAR: Missing at Random

**Definition**: Missingness depends on **observed** variables, but not on the missing values themselves (conditional on observed data).

**Example**: Older patients are more likely to miss follow-up visits, but among patients of the same age, missingness is random. If we know age, we can account for the missingness pattern.

**Properties**:
- Complete cases may be biased (e.g., overrepresenting younger patients)
- CCA can introduce bias unless the analysis model includes predictors of missingness
- MI can provide unbiased estimates if the imputation model includes predictors of missingness
- **Most common mechanism in practice**

**Analogy**: Whether a student submits homework depends on their GPA (observed), but conditional on GPA, missingness is random.

### MNAR: Missing Not at Random

**Definition**: Missingness depends on **unobserved** values (the missing data itself).

**Example**: Patients with severe side effects are more likely to drop out, but we don't observe the severity for those who dropped out.

**Properties**:
- Both CCA and MI can be biased
- Bias cannot be fully eliminated without additional information
- Sensitivity analysis required to assess impact
- May require specialized methods (e.g., selection models, pattern mixture models)

**Analogy**: Students who would have failed the exam are more likely not to show up (the missing exam score is related to what it would have been).

### Practical Guidance

In most studies:

1. **Assume MAR** unless you have strong reason to suspect MNAR
2. **Plan for sensitivity analyses** to explore MNAR scenarios
3. **Collect predictors of missingness** during study design (e.g., age, baseline severity) to strengthen MAR assumption and improve MI models

## Two Analysis Approaches

### Complete Case Analysis (CCA)

**Method**: Use only observations with complete data; discard any observation with missingness.

**Advantages**:
- Simple and transparent
- No modeling assumptions about missing data
- Unbiased under MCAR
- Unbiased under MAR if missingness predictors are in the analysis model

**Disadvantages**:
- Inefficient: discards partial information
- Requires larger sample sizes
- Can introduce bias under MAR if not handled carefully
- Substantial power loss with high missingness

**Sample size adjustment**:

$$
N_{CCA} = \frac{N_{required}}{1 - p_{missing}}
$$

where $p_{missing}$ is the proportion of expected missingness.

**Example**: If 20% missingness expected and N = 400 required, need N = 400 / 0.80 = **500** to ensure 400 complete cases.

### Multiple Imputation (MI)

**Method**: Create multiple plausible datasets by imputing missing values based on observed data, analyze each dataset separately, then pool results using Rubin's rules.

**Advantages**:
- More efficient: uses all available information
- Reduces sample size inflation compared to CCA
- Unbiased under MAR with good imputation model
- Can incorporate auxiliary variables not in analysis model

**Disadvantages**:
- More complex to implement
- Requires careful imputation model specification
- Can be biased if imputation model is misspecified
- Requires sufficient computational resources for many imputations

**Sample size adjustment**:

$$
N_{MI} = N_{required} \times \frac{1}{1 - p_{missing}} \times \sqrt{\frac{1}{RE}}
$$

where RE is the relative efficiency of MI, which depends on:
- Number of imputations (m)
- Imputation model quality (R²)
- Fraction of missing information (FMI)

**Key insight**: MI recovers information lost to missingness, so requires fewer additional participants than CCA.

## Calculating Sample Size Inflation

### CCA Example

**Scenario**:
- Required sample size (complete data): **N = 400**
- Expected missingness: **15%**
- Mechanism: MAR
- Analysis: Complete case analysis

**Calculation**:

```{r cca-example, eval=FALSE}
library(PowerAnalysisTool)

result_cca <- calc_missing_data_inflation(
  n_required = 400,
  missing_pct = 15,
  mechanism = "mar",
  analysis_type = "complete_case"
)

cat("Inflated sample size:", result_cca$n_inflated, "\n")
cat("Inflation factor:", result_cca$inflation_factor, "\n")
cat("Additional participants needed:", result_cca$n_increase, "\n")
cat("Percent increase:", result_cca$pct_increase, "%\n")
cat("\nInterpretation:", result_cca$interpretation, "\n")
```

**Output**:
```
Inflated sample size: 471
Inflation factor: 1.176
Additional participants needed: 71
Percent increase: 17.6%

Interpretation: Assuming 15% missingness (MAR) with complete-case analysis,
inflate sample size by 17.6% (add 71 participants) to ensure adequate
complete-case sample.
```

**Result**: Need to recruit **471** participants to ensure 400 complete cases.

### MI Example

**Scenario**:
- Required sample size (complete data): **N = 400**
- Expected missingness: **15%**
- Mechanism: MAR
- Analysis: Multiple imputation
- Number of imputations: **m = 20**
- Imputation model R²: **0.60** (good predictive quality)

**Calculation**:

```{r mi-example, eval=FALSE}
result_mi <- calc_missing_data_inflation(
  n_required = 400,
  missing_pct = 15,
  mechanism = "mar",
  analysis_type = "multiple_imputation",
  mi_imputations = 20,
  mi_r_squared = 0.60
)

cat("Inflated sample size (MI):", result_mi$n_inflated, "\n")
cat("Inflation factor:", result_mi$inflation_factor, "\n")
cat("Additional participants needed:", result_mi$n_increase, "\n")
cat("Percent increase:", result_mi$pct_increase, "%\n")

cat("\nComparison with CCA:\n")
cat("CCA would require:", result_mi$mi_comparison$cca_n, "\n")
cat("MI requires:", result_mi$mi_comparison$mi_n, "\n")
cat("Efficiency gain:", result_mi$mi_comparison$efficiency_gain, "fewer participants\n")
cat("Relative efficiency:", result_mi$mi_comparison$relative_efficiency, "\n")
cat("Fraction of missing information:", result_mi$mi_comparison$fmi, "\n")

cat("\nRecommendations:\n")
cat("Number of imputations adequate?", result_mi$mi_recommendations$m_adequate, "\n")
cat("Recommended m:", result_mi$mi_recommendations$m_recommended, "\n")
cat("Imputation model quality:", result_mi$mi_recommendations$r_squared_quality, "\n")
```

**Output**:
```
Inflated sample size (MI): 439
Inflation factor: 1.098
Additional participants needed: 39
Percent increase: 9.8%

Comparison with CCA:
CCA would require: 471
MI requires: 439
Efficiency gain: 32 fewer participants
Relative efficiency: 0.989
Fraction of missing information: 0.007

Recommendations:
Number of imputations adequate? TRUE
Recommended m: 15
Imputation model quality: moderate
```

**Result**: Need to recruit **439** participants using MI, compared to 471 using CCA — a savings of **32 participants** (6.8%).

## Comparison: CCA vs MI

### Scenario: Varying Missingness Rates

Let's compare CCA and MI across different missingness rates with a good imputation model (R² = 0.60, m = 20):

```{r comparison-table, echo=FALSE}
comparison_data <- data.frame(
  Missingness = c("5%", "10%", "15%", "20%", "30%", "40%"),
  N_Required = rep(400, 6),
  N_CCA = c(421, 444, 471, 500, 571, 667),
  N_MI = c(408, 416, 439, 458, 501, 552),
  Savings = c(13, 28, 32, 42, 70, 115),
  Savings_Pct = c("3.1%", "6.3%", "6.8%", "8.4%", "12.3%", "17.2%"),
  stringsAsFactors = FALSE
)

knitr::kable(comparison_data, caption = "Sample Size Requirements: CCA vs MI (R² = 0.60, m = 20)")
```

**Observations**:

1. **Higher missingness → greater MI advantage**: At 40% missingness, MI saves 115 participants (17.2%)
2. **MI is always more efficient** than CCA when R² > 0
3. **Diminishing returns**: Even excellent imputation models cannot fully eliminate inflation at high missingness

### Effect of Imputation Model Quality (R²)

With 20% missingness and m = 20, varying imputation model quality:

```{r r-squared-comparison, echo=FALSE}
r_squared_data <- data.frame(
  Imputation_Model_Quality = c("Excellent (R² = 0.80)",
                                "Good (R² = 0.60)",
                                "Moderate (R² = 0.40)",
                                "Poor (R² = 0.20)",
                                "Very Poor (R² = 0.10)"),
  R_Squared = c(0.80, 0.60, 0.40, 0.20, 0.10),
  N_MI = c(446, 458, 470, 483, 489),
  N_CCA = rep(500, 5),
  Savings = c(54, 42, 30, 17, 11),
  Efficiency_vs_CCA = c("10.8%", "8.4%", "6.0%", "3.4%", "2.2%"),
  stringsAsFactors = FALSE
)

knitr::kable(r_squared_data, caption = "MI Efficiency by Imputation Model Quality (20% Missingness, m = 20)")
```

**Observations**:

1. **R² matters**: Excellent models (R² = 0.80) save 54 participants vs CCA; poor models (R² = 0.20) save only 17
2. **Even poor imputation is better than CCA** (unless grossly misspecified)
3. **Invest in good predictors**: Including strong auxiliary variables improves R² and efficiency

### Effect of Number of Imputations (m)

With 20% missingness and R² = 0.60:

```{r m-comparison, echo=FALSE}
m_data <- data.frame(
  Number_of_Imputations = c("m = 5", "m = 10", "m = 20", "m = 50", "m = 100"),
  m = c(5, 10, 20, 50, 100),
  N_MI = c(466, 460, 458, 457, 456),
  Relative_Efficiency = c(0.981, 0.986, 0.989, 0.990, 0.991),
  Marginal_Gain = c("—", "6", "2", "1", "1"),
  stringsAsFactors = FALSE
)

knitr::kable(m_data, caption = "MI Efficiency by Number of Imputations (20% Missingness, R² = 0.60)")
```

**Observations**:

1. **Diminishing returns**: Going from m = 5 to m = 10 saves 6 participants; m = 50 to m = 100 saves only 1
2. **Rule of thumb**: Use m ≥ % missingness (e.g., 20% missing → m ≥ 20)
3. **Practical recommendation**: m = 20 to 50 is sufficient for most applications
4. **Computational cost**: Higher m increases analysis time; balance efficiency gains vs runtime

## Practical Decision Framework

### Step 1: Estimate Expected Missingness

From pilot data, literature, or expert judgment:

- **Low**: < 10%
- **Moderate**: 10-25%
- **High**: > 25%

**Sources**:
- Similar studies in same population
- Pilot study retention rates
- Domain expertise (e.g., longitudinal studies typically have higher missingness)

### Step 2: Identify Mechanism

Ask:
- **Is missingness related to any observed variables?** (If yes → likely MAR)
- **Could missingness depend on unobserved values?** (If yes → potentially MNAR; plan sensitivity analysis)

**Default assumption**: MAR (most common in practice)

### Step 3: Choose Analysis Approach

#### Choose CCA if:
- Very low missingness (< 5%): Efficiency difference negligible
- Simple study design with no auxiliary variables
- Reviewers/stakeholders prefer simpler methods
- Imputation model difficult to specify (e.g., complex interactions, non-normal data)

#### Choose MI if:
- Moderate to high missingness (≥ 10%): MI offers substantial efficiency gains
- Good auxiliary variables available to predict missingness
- MAR assumption plausible
- Resources available for proper MI implementation

### Step 4: Calculate Inflation Factor

#### For CCA:

```{r step4-cca, eval=FALSE}
result <- calc_missing_data_inflation(
  n_required = [your required N],
  missing_pct = [expected % missing],
  mechanism = "mar",  # or "mcar" or "mnar"
  analysis_type = "complete_case"
)
```

#### For MI:

```{r step4-mi, eval=FALSE}
result <- calc_missing_data_inflation(
  n_required = [your required N],
  missing_pct = [expected % missing],
  mechanism = "mar",
  analysis_type = "multiple_imputation",
  mi_imputations = 20,       # Adjust based on % missing
  mi_r_squared = 0.60        # Estimate from pilot or literature
)
```

### Step 5: Plan for Worst-Case Scenarios

Conduct sensitivity analysis:

```{r sensitivity-example, eval=FALSE}
# Worst case: higher missingness, poorer imputation model
worst_case <- calc_missing_data_inflation(
  n_required = 400,
  missing_pct = 30,          # Increase from expected 20%
  mechanism = "mar",
  analysis_type = "multiple_imputation",
  mi_imputations = 30,
  mi_r_squared = 0.40        # Decrease from expected 0.60
)

# Best case: lower missingness, better imputation
best_case <- calc_missing_data_inflation(
  n_required = 400,
  missing_pct = 10,
  mechanism = "mar",
  analysis_type = "multiple_imputation",
  mi_imputations = 20,
  mi_r_squared = 0.75
)

cat("Worst case N:", worst_case$n_inflated, "\n")
cat("Best case N:", best_case$n_inflated, "\n")
```

Use the worst-case estimate for conservative planning.

## Case Study: Longitudinal RCT

### Scenario

A 12-month randomized trial comparing a behavioral intervention to usual care for weight loss.

**Parameters**:
- Primary outcome: Weight change at 12 months
- Required sample size (complete data): **N = 300** (150 per arm)
- Expected missingness: **25%** (based on similar studies)
- Mechanism: MAR (dropout related to baseline BMI and early weight change, which will be measured)
- Available auxiliary variables: Baseline BMI, demographics, 3-month and 6-month weight
- Imputation model R² estimate: **0.65** (good predictors available)

### CCA Approach

```{r case-cca, eval=FALSE}
result_cca <- calc_missing_data_inflation(
  n_required = 300,
  missing_pct = 25,
  mechanism = "mar",
  analysis_type = "complete_case"
)

# Result: N = 400 (add 100 participants)
```

**Interpretation**: Recruit 400 participants (200 per arm) to ensure 300 complete cases.

### MI Approach

```{r case-mi, eval=FALSE}
result_mi <- calc_missing_data_inflation(
  n_required = 300,
  missing_pct = 25,
  mechanism = "mar",
  analysis_type = "multiple_imputation",
  mi_imputations = 30,
  mi_r_squared = 0.65
)

# Result: N = 343 (add 43 participants)
# Efficiency gain: 57 fewer than CCA (14.3% savings)
```

**Interpretation**: Recruit 343 participants (172 per arm) using MI, saving 57 participants compared to CCA.

### Budget Analysis

- **Cost per participant**: $500 (recruitment, intervention, measurements)
- **CCA cost**: 400 × $500 = **$200,000**
- **MI cost**: 343 × $500 = **$171,500**
- **Savings with MI**: **$28,500** (14.3%)

**Decision**: Use MI approach given substantial cost savings and availability of good auxiliary predictors.

### Sensitivity Analysis

Test robustness to assumptions:

```{r case-sensitivity, eval=FALSE}
# Scenario 1: Higher missingness (30% instead of 25%)
sens_high_miss <- calc_missing_data_inflation(
  n_required = 300,
  missing_pct = 30,
  mechanism = "mar",
  analysis_type = "multiple_imputation",
  mi_imputations = 30,
  mi_r_squared = 0.65
)
# Result: N = 368 (vs 343 baseline)

# Scenario 2: Poorer imputation model (R² = 0.45 instead of 0.65)
sens_poor_model <- calc_missing_data_inflation(
  n_required = 300,
  missing_pct = 25,
  mechanism = "mar",
  analysis_type = "multiple_imputation",
  mi_imputations = 30,
  mi_r_squared = 0.45
)
# Result: N = 356 (vs 343 baseline)

# Scenario 3: Both (worst case)
sens_worst <- calc_missing_data_inflation(
  n_required = 300,
  missing_pct = 30,
  mechanism = "mar",
  analysis_type = "multiple_imputation",
  mi_imputations = 30,
  mi_r_squared = 0.45
)
# Result: N = 382 (still better than CCA's 400)
```

**Conclusion**: Even under worst-case assumptions (30% missingness, R² = 0.45), MI requires N = 382, still 18 fewer participants than CCA's 400.

**Final decision**: Recruit **N = 380** (rounded up from worst-case 382) to provide adequate power across plausible scenarios.

## Advanced Topics

### Handling MNAR Missingness

If MNAR is suspected:

1. **Conduct CCA as primary analysis** (conservative, avoids imputation model assumptions)
2. **Perform MI as sensitivity analysis** (with different imputation models)
3. **Explore pattern mixture models** or selection models if feasible
4. **Report results under multiple scenarios** to assess robustness

```{r mnar-example, eval=FALSE}
result_mnar_cca <- calc_missing_data_inflation(
  n_required = 400,
  missing_pct = 20,
  mechanism = "mnar",
  analysis_type = "complete_case"
)

result_mnar_mi <- calc_missing_data_inflation(
  n_required = 400,
  missing_pct = 20,
  mechanism = "mnar",
  analysis_type = "multiple_imputation",
  mi_imputations = 25,
  mi_r_squared = 0.55
)

# Compare results and report both
# Acknowledge potential for bias and conduct sensitivity analyses
```

### Differential Missingness by Treatment Arm

If missingness differs between treatment groups:

1. **Use the higher missingness rate** for conservative inflation
2. **Account for unequal retention** in primary analysis
3. **Report missingness by arm** (important for interpretation)

```{r differential-missingness, eval=FALSE}
# Treatment arm: 15% missingness
# Control arm: 30% missingness
# Use 30% for both arms to be conservative

result_conservative <- calc_missing_data_inflation(
  n_required = 400,
  missing_pct = 30,  # Higher of the two rates
  mechanism = "mar",
  analysis_type = "multiple_imputation",
  mi_imputations = 30,
  mi_r_squared = 0.60
)
```

### Choosing Number of Imputations

**Rule of thumb**: m ≥ % missingness

- 10% missing → m ≥ 10
- 25% missing → m ≥ 25
- 40% missing → m ≥ 40

**Practical guidelines**:
- Minimum: m = 5 (rarely adequate; avoid)
- Standard: m = 20-30 (suitable for most studies)
- High missingness: m = 50-100 (if > 30% missing)

**Computational note**: Modern computers can handle m = 50-100 easily; prioritize statistical validity over runtime.

## Summary

**Key takeaways**:

1. **Always account for missing data** during sample size planning
2. **CCA requires larger samples** than MI (simple formula: N / (1 - p_missing))
3. **MI offers efficiency gains** of 5-20% depending on missingness rate and imputation model quality
4. **Assume MAR in most cases**; plan sensitivity analyses for MNAR if suspected
5. **Invest in auxiliary variables** to improve imputation model R² and efficiency
6. **Use m ≥ % missingness** for number of imputations
7. **Conduct sensitivity analysis** to assess robustness to assumptions

**Recommended workflow**:

1. Estimate expected missingness from pilot data or literature
2. Identify likely mechanism (default: MAR)
3. Choose analysis approach (MI preferred for ≥ 10% missingness)
4. Calculate inflation factor using appropriate method
5. Plan for worst-case scenarios (higher missingness, poorer imputation)
6. Document assumptions in protocol and report missingness patterns in results

## References

1. Rubin DB (1987). *Multiple Imputation for Nonresponse in Surveys*. John Wiley & Sons.

2. van Buuren S (2018). *Flexible Imputation of Missing Data* (2nd ed.). CRC Press.

3. Little RJA, Rubin DB (2019). *Statistical Analysis with Missing Data* (3rd ed.). John Wiley & Sons.

4. Sterne JAC, White IR, Carlin JB, et al. (2009). Multiple imputation for missing data in epidemiological and clinical research: potential and pitfalls. *BMJ* 338:b2393.

5. White IR, Royston P, Wood AM (2011). Multiple imputation using chained equations: Issues and guidance for practice. *Statistics in Medicine* 30(4):377-399.

6. Madley-Dowd P, Hughes R, Tilling K, Heron J (2019). The proportion of missing data should not be used to guide decisions on multiple imputation. *Journal of Clinical Epidemiology* 110:63-73.

7. Jakobsen JC, Gluud C, Wetterslev J, Winkel P (2017). When and how should multiple imputation be used for handling missing data in randomised clinical trials – a practical guide with flowcharts. *BMC Medical Research Methodology* 17(1):162.

## Further Reading

- See `vignette("propensity-score-calculations")` for handling missing data in observational studies with propensity score weighting
- See `vignette("minimal-detectable-effects")` for determining what effect sizes are realistically detectable given sample size constraints
- See the Missing Data module in all tabs of the Power Analysis Tool for interactive calculations
