---
title: "`r if(params$reference_proportion == 0) 'Single Proportion Analysis: Rare Event Detection' else 'Single Proportion Power Analysis'`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    df_print: kable
    keep_tex: false
fontsize: 12pt
geometry: margin=1in
always_allow_html: yes
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{threeparttable}
params:
  tabset: 'Sample Size'
  expected_proportion: 1
  reference_proportion: 0
  sample_size: 230
  power: 0.8
  discon: 0.10
  adj_n: 250
  progress: NA
---

```{r, include=FALSE}
options(tinytex.verbose = F)
```

```{r setup, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pwr)
library(binom)
library(kableExtra)
```

## About this analysis
This analysis performs power and sample size calculations for single proportion hypothesis testing using Cohen's arcsine transformation method. This statistical approach tests whether an observed proportion differs significantly from a reference value.

The method uses the arcsine transformation of proportions, which provides accurate power calculations across a wide range of proportions. This is the standard approach recommended in Cohen (1988) "Statistical Power Analysis for the Behavioral Sciences" and is implemented using the pwr.p.test function in R.

`r if(params$reference_proportion == 0) {
  paste0("This analysis is configured for rare event detection (reference proportion = 0%), useful for post-marketing surveillance, safety monitoring, and quality control where the goal is to detect if any events occur.")
} else {
  paste0("This analysis compares an expected proportion against a reference rate of ", params$reference_proportion, "%, useful for testing against historical controls, benchmarks, or standards of care.")
}`

Reference: Cohen J. Statistical Power Analysis for the Behavioral Sciences. 2nd ed. Routledge; 1988.

## Results of this analysis
`r if(params$reference_proportion == 0) {
  paste0("With an expected event rate of ", format(params$expected_proportion, digits=2, nsmall=2), "% (1 in ", format(100/params$expected_proportion, digits=1), "), ", format(ceiling(params$sample_size), digits=1, nsmall=0), " participants would be needed to achieve ", format(params$power*100, digits=2, nsmall=0), "% power for detecting at least one event. Accounting for a possible withdrawal or discontinuation rate of ", format(params$discon*100, digits=1), "%, the target sample size is ", format(ceiling(params$sample_size * (1+params$discon)), digits=1), " participants.")
} else {
  paste0("To detect a difference from the reference proportion of ", params$reference_proportion, "% to an expected proportion of ", params$expected_proportion, "%, ", format(ceiling(params$sample_size), digits=1, nsmall=0), " participants are required to achieve ", format(params$power*100, digits=2, nsmall=0), "% statistical power. Accounting for a possible withdrawal or discontinuation rate of ", format(params$discon*100, digits=1), "%, the target sample size is ", format(ceiling(params$sample_size * (1+params$discon)), digits=1), " participants.")
}`

\newpage
### Figure 1. Estimated power for the given conditions at different sample sizes.
```{r figure1, echo=FALSE}
# Increment the progress bar, and update the detail text.
params$progress$inc(1/3, detail = "Making Figure 2")

p.out <- if(params$tabset == 'Power'){
    pwr.p.test(sig.level=0.05, power=NULL, h = ES.h(params$expected_proportion/100, params$reference_proportion/100), alt="greater", n = params$sample_size)
  } else {
    pwr.p.test(sig.level=0.05, power=params$power, h = ES.h(params$expected_proportion/100, params$reference_proportion/100), alt="greater", n = NULL)
  }

plot(p.out)
```

\newpage
### Table 1. Event rate accuracy with 95% confidence interval limits
If `r format(ceiling(params$sample_size), digits=1, nsmall=0)` participants are included, the event rate would be estimated to an accuracy shown in the table below:


```{r table1, echo=FALSE, results='asis'}
# Increment the progress bar, and update the detail text.
params$progress$inc(1/4, detail = "Creating Table 1")

sequence <- unique(c(seq(0, 5), seq(10, 25, by=5), seq(50, min(round(params$sample_size, 0), 1000), by=50), seq(min(round(params$sample_size, 0), 1000), min(round(params$sample_size, 0), 10000), by=1000)))
bb <- lapply(sequence, function(n) {
  binom.confint(n, params$sample_size, conf.level = 0.95, methods = "exact")
})

table <- do.call(rbind, bb)
table$length <- table$upper - table$lower
var <- c("mean", "lower", "upper", "length")
for (i in var) {
  table[, i] <- round(table[, i]*100, 1)
}

table <- table[, c(2,4:7)]
kable(table, align='lrrrr', 
      col.names = c('Number of Events Observed', "Event Rate$^1$", "Lower Limit$^2$", "Upper Limit$^2$",'Length'), 
      format = "latex", booktabs=T, escape=F) %>%
  kable_styling(full_width = FALSE, position="center", fixed_thead = T) %>%
  add_header_above(c(" ", " ", "95% Confidence Interval" = 2, " ")) %>%
  footnote(number = c("Event rate (%) is estimated as a crude rate, defined as the number of participants exposed and experiencing the event of interest divided by the total number of participants. ",
                      "Confidence Interval (%) based on exact Clopper-Pearson exact method for one proportion.")
           , threeparttable = T)
params$progress$inc(1/5, detail = "Writing PDF File (May take a few minutes)")
```